---
title: "Figure 4 pilot:  taxonomy benchmark"
author: "Kim So-Yeon"
date: "2026-07-27"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)
options(stringsAsFactors = FALSE)
```

# 1. Brief analysis plan

This Rmd processes one completed ITSdetector run and compares two taxonomy versions using the same ASV abundance table.

- UNITE before BLAST: `taxonomy_normalized.tsv`
- UNITE+BLAST after refinement: `taxonomy_blast.tsv` with normalized rank columns
- Species names are harmonized before comparison with the expected mock composition.
- Main detection rule: `count > 0`
- PBS negative controls are retained in composition plots but excluded from ground-truth accuracy summaries.

# 2. Load packages and paths


## 1) pacakges and paths
```{r load-packages-paths}

library(tidyverse)
library(qiime2R)
library(ggplot2)
library(ggh4x)
library(vegan)
library(randomForest)
library(tidyverse)
library(patchwork)
library(scales)

```


```{r}

project_dir <- "/data/home2/ksy/260724_GutMyco_Fungal_Mock_Paper"
rproj_dir <- file.path(project_dir, "Rproj")

analysis_dir <- file.path(rproj_dir, "Output", "fig4_pilot_all4_corrected4")
table_dir <- file.path(analysis_dir, "tables")
figure_dir <- file.path(analysis_dir, "figures")

dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
```


## 2) Raw data path
```{r}
run_tbl <- tibble::tribble(
  ~run_id,                       ~preprocess_method, ~read_strategy, ~preprocess_dir, ~read_dir,
  "fig4_q30_PE_itsxpress_unite_blast_top1_rescue", "ITSxpress",        "PE",           "itsxpress",     "paired",
  "fig4_q30_SE_itsxpress_unite_blast_top1_rescue", "ITSxpress",        "SE",           "itsxpress",     "single",
  "fig4_q30_PE_cutadapt_unite_blast_top1_rescue",  "Cutadapt",         "PE",           "cutadapt",      "paired",
  "fig4_q30_SE_cutadapt_unite_blast_top1_rescue",  "Cutadapt",         "SE",           "cutadapt",      "single"
) |>
  dplyr::mutate(
    run_dir = file.path(project_dir, "Output", .data$run_id),
    pipeline = "ITSdetector",
    Paired = .data$read_strategy,
    Trimm = .data$preprocess_method,
    table_qza = file.path(.data$run_dir, "dada2", .data$preprocess_dir, .data$read_dir, "table.qza"),
    repseq_qza = file.path(.data$run_dir, "dada2", .data$preprocess_dir, .data$read_dir, "repseq.qza"),
    denoising_stats_qza = file.path(.data$run_dir, "dada2", .data$preprocess_dir, .data$read_dir, "denoising_stats.qza"),
    taxonomy_unite_normalized = file.path(.data$run_dir, "taxonomy_blast", .data$preprocess_dir, .data$read_dir, "unite_ver10_99", "taxonomy_normalized.tsv"),
    taxonomy_blast_final = file.path(.data$run_dir, "taxonomy_blast", .data$preprocess_dir, .data$read_dir, "unite_ver10_99", "taxonomy_blast.tsv"),
    taxonomy_blast_qiime = file.path(.data$run_dir, "taxonomy_blast", .data$preprocess_dir, .data$read_dir, "unite_ver10_99", "taxonomy_blast_qiime.tsv"),
    taxonomy_custom = file.path(run_dir, "taxonomy_blast", preprocess_dir, read_dir, "CustomDB", "taxonomy.tsv"),
    taxonomy_custom_evidence = file.path(run_dir, "taxonomy_blast", preprocess_dir, read_dir, "CustomDB", "taxonomy_customdb_evidence.tsv"),

    blast_evidence = file.path(.data$run_dir, "taxonomy_blast", .data$preprocess_dir, .data$read_dir, "unite_ver10_99", "taxonomy_blast_evidence.tsv"),
    blast_report = file.path(.data$run_dir, "taxonomy_blast", .data$preprocess_dir, .data$read_dir, "unite_ver10_99", "taxonomy_blast_report.tsv"),
    cleanup_qc_summary = file.path(.data$run_dir, "cleanup_qc", "cleanup_qc_summary.tsv")
  )


sample_metadata_path <- file.path(project_dir, "Input", "metadata_mock_stool_and_mock_260724.tsv")

path_check <- run_tbl |>
  dplyr::select(
    run_id,
    table_qza,
    taxonomy_unite_normalized,
    taxonomy_blast_final,
    taxonomy_custom,
    taxonomy_custom_evidence
  ) |>
  tidyr::pivot_longer(
    cols = -run_id,
    names_to = "file_key",
    values_to = "path"
  ) |>
  dplyr::bind_rows(
    tibble::tibble(
      run_id = "metadata",
      file_key = "sample_metadata",
      path = sample_metadata_path
    )
  ) |>
  dplyr::mutate(exists = file.exists(path))

print(path_check)

missing_required <- path_check |>
  dplyr::filter(!exists)

missing_required
# if (nrow(missing_required) > 0) {
#   stop("Missing required input files:\n", paste(missing_required$path, collapse = "\n"))
# }
```

모든 파일 다 인식 완료 

# 3. Define expected mock composition and plotting style

## 1) Mock community (ATCC)
```{r expected-style}
expected_species <- c(
  "Aspergillus_fumigatus",
  "Cryptococcus_neoformans",
  "Trichophyton_interdigitale",
  "Penicillium_chrysogenum",
  "Fusarium_keratoplasticum",
  "Candida_albicans",
  "Nakaseomyces_glabratus",
  "Malassezia_globosa",
  "Saccharomyces_cerevisiae",
  "Cutaneotrichosporon_dermatis"
)

expected_genus <- c(
  "Aspergillus", "Cryptococcus", "Trichophyton", "Penicillium", "Fusarium",
  "Candida", "Nakaseomyces", "Malassezia", "Saccharomyces", "Cutaneotrichosporon"
)

expected_truth <- dplyr::bind_rows(
  tibble::tibble(taxonomic_rank = "Species", taxon = expected_species, expected_abundance = 0.1),
  tibble::tibble(taxonomic_rank = "Genus", taxon = expected_genus, expected_abundance = 0.1)
)


readr::write_tsv(expected_truth, file.path(table_dir, "expected_mock_composition.tsv"))

```

## 2) taxa color
```{r}
tax_col <- c(
  "#E41A1CFF", "#377EB8FF", "#4DAF4AFF", "#984EA3FF", "#FF7F00FF",
  "#FFFF33FF", "#A65628FF", "#f55cacFF", "#E6AB02FF", "#1B9E77FF", "#999999FF"
)

met_col <- c(
  "#66C2A5", "#FC8D62", "#8DA0CB", "#E78AC3",
  "#A6D854", "#FFD92F", "#E5C494", "#B3B3B3"
)
```

## 3) figure theme  
```{r}
fig_theme <- function(base_size = 12) {
  ggplot2::theme_classic(base_size = base_size) +
    ggplot2::theme(
      panel.background = ggplot2::element_blank(),
      plot.background = ggplot2::element_blank(),
      panel.grid = ggplot2::element_blank(),
      axis.line = ggplot2::element_line(color = "black", linewidth = 0.5),
      axis.ticks = ggplot2::element_line(color = "black", linewidth = 0.5),
      axis.text.x.bottom = ggplot2::element_text(angle = 45, hjust = 1.1, vjust = 0.9),
      legend.title = ggplot2::element_blank(),
      strip.background = ggplot2::element_blank(),
      strip.text = ggplot2::element_text(face = "bold")
    )
}
```

# 4. Helper functions

```{r helper-functions}
clean_taxon_value <- function(x) {
  x |>
    as.character() |>
    stringr::str_replace_all("^.__", "") |>
    stringr::str_replace_all(" ", "_") |>
    stringr::str_replace_all(";.*$", "") |>
    stringr::str_replace_all("^_+|_+$", "") |>
    dplyr::na_if("")
}

harmonize_taxon <- function(x, rank = "Species") {
  x1 <- clean_taxon_value(x)

  dplyr::case_when(
    rank == "Species" & x1 %in% c("Candida_glabrata", "Nakaseomyces_glabrata") ~ "Nakaseomyces_glabratus",
    # rank == "Species" & x1 %in% c("Cryptococcus_neoformans_var_grubii", "Cryptococcus_neoformans_var._grubii") ~ "Cryptococcus_neoformans",
    rank == "Species" & x1 == "Penicillium notatum" ~ "Penicillium_chrysogenum",
    # rank == "Species" & x1 == "Trichophyton_mentagrophytes" ~ "Trichophyton_interdigitale",
    TRUE ~ x1
  )
}

parse_qiime_taxon <- function(taxon_string, rank_index) {
  split_mat <- stringr::str_split_fixed(taxon_string, ";", 7)
  clean_taxon_value(split_mat[, rank_index])
}

read_taxonomy_table <- function(path, taxonomy_method_label) {
  tax <- readr::read_tsv(path, show_col_types = FALSE)

  if ("Feature ID" %in% colnames(tax)) {
    tax <- dplyr::rename(tax, feature_id = `Feature ID`)
  }

  if ("ASV" %in% colnames(tax)) {
    tax <- dplyr::rename(tax, feature_id = ASV)
  }

  if (!"feature_id" %in% colnames(tax)) {
    stop("Taxonomy table has no Feature ID or ASV column: ", path)
  }

  if (!"Genus" %in% colnames(tax) && "Taxon" %in% colnames(tax)) {
    tax <- dplyr::mutate(tax, Genus = parse_qiime_taxon(.data$Taxon, 6))
  }

  if (!"Species" %in% colnames(tax) && "Taxon" %in% colnames(tax)) {
    tax <- dplyr::mutate(tax, Species = parse_qiime_taxon(.data$Taxon, 7))
  }

  if (!all(c("Genus", "Species") %in% colnames(tax))) {
    stop("Taxonomy table must contain Genus and Species columns or a parsable Taxon column: ", path)
  }

  if (!"Taxon" %in% colnames(tax)) {
    tax <- dplyr::mutate(tax, Taxon = NA_character_)
  }

  tax |>
    dplyr::transmute(
      feature_id = as.character(.data$feature_id),
      taxonomy_method = taxonomy_method_label,
      Genus = harmonize_taxon(.data$Genus, rank = "Genus"),
      Species = harmonize_taxon(.data$Species, rank = "Species"),
      taxon_string = as.character(.data$Taxon)
    ) |>
    dplyr::mutate(
      Genus = dplyr::if_else(is.na(.data$Genus) | stringr::str_detect(.data$Genus, "_g$"), "Unclassified", .data$Genus),
      Species = dplyr::if_else(is.na(.data$Species) | stringr::str_detect(.data$Species, "_g$|_s$"), "Unclassified", .data$Species)
    )
}

standardize_metadata <- function(metadata_df) {
  meta <- metadata_df

  if ("sample" %in% colnames(meta) && !"sample_id" %in% colnames(meta)) {
    meta <- dplyr::rename(meta, sample_id = sample)
  }

  if ("kit_code" %in% colnames(meta) && !"kit" %in% colnames(meta)) {
    meta <- dplyr::rename(meta, kit = kit_code)
  }

  meta |>
    dplyr::mutate(
      sample_id = as.character(.data$sample_id),
      sample_id = stringr::str_replace(.data$sample_id, "^X", ""),
      DNA_kit = dplyr::recode(
        as.character(.data$kit),
        "D" = "DNeasy",
        "Q" = "QIAamp",
        "Z" = "Zymo",
        .default = as.character(.data$kit)
      ),
      replicate = as.character(.data$replicate),
      matrix_type = as.character(.data$matrix_type)
    ) |>
    dplyr::select(.data$sample_id, .data$DNA_kit, .data$replicate, .data$matrix_type, dplyr::everything()) |>
    dplyr::distinct(.data$sample_id, .keep_all = TRUE)
}

feature_table_to_long <- function(feature_table_obj, taxonomy_feature_ids) {
  ft <- as.data.frame(feature_table_obj)
  row_match <- sum(rownames(ft) %in% taxonomy_feature_ids)
  col_match <- sum(colnames(ft) %in% taxonomy_feature_ids)

  if (col_match > row_match) {
    ft <- t(ft) |> as.data.frame()
  }

  ft |>
    tibble::rownames_to_column("feature_id") |>
    tidyr::pivot_longer(
      cols = - .data$feature_id,
      names_to = "sample_id",
      values_to = "count"
    ) |>
    dplyr::mutate(
      sample_id = stringr::str_replace(as.character(.data$sample_id), "^X", ""),
      count = as.numeric(.data$count)
    )
}

calc_bray <- function(observed, expected) {
  if (sum(observed, na.rm = TRUE) + sum(expected, na.rm = TRUE) == 0) {
    return(NA_real_)
  }
  as.numeric(vegan::vegdist(rbind(observed, expected), method = "bray")[1])
}

calc_r2 <- function(observed, expected) {
  if (stats::sd(observed, na.rm = TRUE) == 0 || stats::sd(expected, na.rm = TRUE) == 0) {
    return(NA_real_)
  }
  stats::cor(observed, expected, use = "complete.obs")^2
}

calc_metrics <- function(df, threshold_value, threshold_label) {
  df |>
    dplyr::mutate(
      detected = .data$relative_abundance > threshold_value,
      expected_detected = .data$expected_abundance > 0
    ) |>
    dplyr::filter(.data$matrix_type %in% c("mock_only", "mock_plus_stool")) |>
    dplyr::group_by(
      .data$run_id, .data$sample_id, .data$DNA_kit, .data$replicate, .data$matrix_type,
      .data$pipeline, .data$preprocess_method, .data$read_strategy,
      .data$taxonomy_method, .data$BLAST, .data$Paired, .data$Trimm, .data$taxonomic_rank
    ) |>
    dplyr::summarise(
      TP = sum(.data$detected & .data$expected_detected, na.rm = TRUE),
      FP = sum(.data$detected & !.data$expected_detected, na.rm = TRUE),
      FN = sum(!.data$detected & .data$expected_detected, na.rm = TRUE),
      precision = dplyr::if_else(.data$TP + .data$FP > 0, .data$TP / (.data$TP + .data$FP), NA_real_),
      recall = dplyr::if_else(.data$TP + .data$FN > 0, .data$TP / (.data$TP + .data$FN), NA_real_),
      F1 = dplyr::if_else(!is.na(.data$precision) & !is.na(.data$recall) & (.data$precision + .data$recall) > 0,
                          2 * .data$precision * .data$recall / (.data$precision + .data$recall), NA_real_),
      false_discovery_rate = dplyr::if_else(.data$TP + .data$FP > 0, .data$FP / (.data$TP + .data$FP), NA_real_),
      Bray_Curtis = calc_bray(.data$relative_abundance, .data$expected_abundance),
      R2 = calc_r2(.data$relative_abundance, .data$expected_abundance),
      .groups = "drop"
    ) |>
    dplyr::mutate(detection_threshold = threshold_label)
}

make_rf_importance <- function(meta_df, outcome_col) {
  predictors <- c("DNA_kit", "BLAST", "Paired", "Trimm", "matrix_type")
  active_predictors <- predictors[purrr::map_lgl(predictors, ~ dplyr::n_distinct(meta_df[[.x]], na.rm = TRUE) > 1)]

  rf_df <- meta_df |>
    dplyr::filter(!is.na(.data[[outcome_col]])) |>
    dplyr::select(dplyr::all_of(c(outcome_col, active_predictors))) |>
    tidyr::drop_na() |>
    dplyr::mutate(dplyr::across(dplyr::all_of(active_predictors), as.factor))

  if (length(active_predictors) == 0 || nrow(rf_df) < 10) {
    return(list(model = NULL, importance = tibble::tibble()))
  }

  set.seed(123)
  rf_model <- randomForest::randomForest(
    formula = stats::as.formula(paste(outcome_col, "~", paste(active_predictors, collapse = " + "))),
    data = rf_df,
    importance = TRUE
  )

  rf_imp <- randomForest::importance(rf_model, type = 1) |>
    as.data.frame() |>
    tibble::rownames_to_column("Variable") |>
    dplyr::rename(IncMSE = `%IncMSE`) |>
    dplyr::mutate(outcome = outcome_col)

  list(model = rf_model, importance = rf_imp)
}

plot_box_jitter <- function(data, x_var, y_var, fill_var, fill_colors, y_lab = NULL) {
  ggplot2::ggplot(
    data,
    ggplot2::aes(x = .data[[x_var]], y = .data[[y_var]], fill = .data[[fill_var]])
  ) +
    ggplot2::geom_boxplot(width = 0.6, color = "black", linewidth = 0.4, outlier.shape = NA, alpha = 0.6) +
    ggplot2::geom_jitter(width = 0.15, size = 2, shape = 21, stroke = 0.4, color = "black", fill = "white") +
    ggplot2::scale_fill_manual(values = fill_colors) +
    ggplot2::labs(x = NULL, y = y_lab %||% y_var) +
    fig_theme(base_size = 12)
}



plot_bar <-
  function(data,
           x_var,
           y_var,
           fill_var,
           fill_colors,
           y_lab = NULL) {
    plot_df <- data |>
      dplyr::group_by(metric,
                      Trimm,
                      Paired,
                      matrix_type,
                      DNA_kit,
                      dplyr::across(dplyr::all_of(c(x_var, fill_var)))) |>
      dplyr::summarise(
        mean = mean(.data[[y_var]], na.rm = TRUE),
        sd = sd(.data[[y_var]], na.rm = TRUE),
        n = dplyr::n(),
        se = sd / sqrt(n),
        .groups = "drop"
      )
    
    ggplot2::ggplot(plot_df,
                    ggplot2::aes(x = .data[[x_var]],
                                 y = mean,
                                 fill = .data[[fill_var]])) +
      ggplot2::geom_col(width = 0.7,
                        color = "black",
                        linewidth = 0.4) +
      ggplot2::geom_errorbar(
        ggplot2::aes(ymin = mean - se,
                     ymax = mean + se),
        width = 0.2,
        linewidth = 0.5
      ) +
      ggplot2::scale_fill_manual(values = fill_colors) +
      ggplot2::labs(x = NULL,
                    y = y_lab %||% y_var) +
      fig_theme(base_size = 12)
  }

`%||%` <- function(x, y) if (is.null(x)) y else x
```

# 5. Load and prepare data

```{r prepare-data}
sample_metadata <-
  readr::read_tsv(sample_metadata_path, show_col_types = FALSE) |>
  standardize_metadata()

read_one_run <- function(run_row) {
  feature_table_i <- qiime2R::read_qza(run_row$table_qza)$data
  
  taxonomy_unite_i <-
    read_taxonomy_table(run_row$taxonomy_unite_normalized, "UNITE") |>
    dplyr::mutate(run_id = run_row$run_id)
  
  taxonomy_blast_i <-
    read_taxonomy_table(run_row$taxonomy_blast_final, "UNITE+RefSeq BLAST") |>
    dplyr::mutate(run_id = run_row$run_id)
  
  taxonomy_custom_i <-
    read_taxonomy_table(run_row$taxonomy_custom, "Mock type-strain BLAST") |>
    dplyr::mutate(run_id = run_row$run_id)
  
  taxonomy_all_i <- dplyr::bind_rows(taxonomy_unite_i,
                                     taxonomy_blast_i,
                                     taxonomy_custom_i) |>
    dplyr::mutate(
      preprocess_method = run_row$preprocess_method,
      read_strategy = run_row$read_strategy,
      Paired = run_row$Paired,
      Trimm = run_row$Trimm
    )
  
  feature_long_i <-
    feature_table_to_long(feature_table_i, unique(taxonomy_all_i$feature_id)) |>
    dplyr::left_join(sample_metadata, by = "sample_id") |>
    dplyr::mutate(
      run_id = run_row$run_id,
      pipeline = run_row$pipeline,
      preprocess_method = run_row$preprocess_method,
      read_strategy = run_row$read_strategy,
      Paired = run_row$Paired,
      Trimm = run_row$Trimm
    )
  
  sample_depth_i <- feature_long_i |>
    dplyr::group_by(run_id, sample_id, preprocess_method, read_strategy) |>
    dplyr::summarise(total_reads = sum(count, na.rm = TRUE),
                     .groups = "drop")
  
  list(
    taxonomy_all = taxonomy_all_i,
    feature_long = feature_long_i,
    sample_depth = sample_depth_i
  )
}

loaded_runs <- run_tbl |>
  split(seq_len(nrow(run_tbl))) |>
  purrr::map(read_one_run)

taxonomy_all <- loaded_runs |>
  purrr::map("taxonomy_all") |>
  dplyr::bind_rows()

feature_long <- loaded_runs |>
  purrr::map("feature_long") |>
  dplyr::bind_rows()

sample_depth <- loaded_runs |>
  purrr::map("sample_depth") |>
  dplyr::bind_rows()


# readr::write_tsv(taxonomy_all,
#                  file.path(table_dir, "taxonomy_all_runs_used.tsv"))
# readr::write_tsv(feature_long,
#                  file.path(table_dir, "feature_table_asv_long_all_runs.tsv"))
# readr::write_tsv(sample_depth,
#                  file.path(table_dir, "sample_depth_all_runs.tsv"))
```


```{r}
# Sun Jul 26 00:14:05 2026 ------------------------------

taxonomy_all <-   readr::read_tsv("./Output/fig4_pilot_all4_corrected4/tables/taxonomy_all_runs_used.tsv")
feature_long <-   readr::read_tsv("./Output/fig4_pilot_all4_corrected4/tables/feature_table_asv_long_all_runs.tsv")
sample_depth <-   readr::read_tsv("./Output/fig4_pilot_all4_corrected4/tables/sample_depth_all_runs.tsv")


```


# 6. Create benchmark long table

```{r benchmark-long}
taxa_abundance <- feature_long |>
  dplyr::select(
    run_id,
    feature_id,
    sample_id,
    count,
    DNA_kit,
    replicate,
    matrix_type,
    pipeline,
    preprocess_method,
    read_strategy,
    Paired,
    Trimm
  ) |>
  dplyr::filter(matrix_type %in% c("mock_only")) |>
  dplyr::left_join(
    taxonomy_all,
    by = c(
      "run_id",
      "feature_id",
      "preprocess_method",
      "read_strategy",
      "Paired",
      "Trimm"
    )
  ) |>
  dplyr::mutate(
    BLAST = dplyr::case_when(
      taxonomy_method == "UNITE" ~ "No_BLAST",
      taxonomy_method == "UNITE+RefSeq BLAST" ~ "RefSeq_BLAST",
      taxonomy_method == "Mock type-strain BLAST" ~ "Mock_type_reference_BLAST",
      TRUE ~ taxonomy_method
    ),
    Genus = dplyr::if_else(is.na(Genus) |
                             Genus == "", "Unclassified", Genus),
    Species = dplyr::if_else(is.na(Species) |
                               Species == "", "Unclassified", Species)
  ) |>
  tidyr::pivot_longer(
    cols = c(Genus, Species),
    names_to = "taxonomic_rank",
    values_to = "taxon"
  ) |>
  dplyr::mutate(taxon = dplyr::if_else(is.na(taxon) |
                                         taxon == "", "Unclassified", taxon)) |>
  dplyr::group_by(
    run_id,
    sample_id,
    DNA_kit,
    replicate,
    matrix_type,
    pipeline,
    preprocess_method,
    read_strategy,
    taxonomy_method,
    BLAST,
    Paired,
    Trimm,
    taxonomic_rank,
    taxon
  ) |>
  dplyr::summarise(count = sum(count, na.rm = TRUE), .groups = "drop") |>
  dplyr::left_join(sample_depth,
                   by = c("run_id", "sample_id", "preprocess_method", "read_strategy")) |>
  dplyr::mutate(
    total_reads = tidyr::replace_na(total_reads, 0),
    relative_abundance = dplyr::if_else(total_reads > 0, count / total_reads, 0)
  )

sample_method_grid <- taxa_abundance |>
  dplyr::distinct(
    run_id,
    sample_id,
    DNA_kit,
    replicate,
    matrix_type,
    pipeline,
    preprocess_method,
    read_strategy,
    taxonomy_method,
    BLAST,
    Paired,
    Trimm
  )

taxon_universe <- dplyr::bind_rows(
  taxa_abundance |> dplyr::distinct(taxonomic_rank, taxon),
  expected_truth |> dplyr::distinct(taxonomic_rank, taxon)
) |>
  dplyr::distinct(taxonomic_rank, taxon)

benchmark_long <- sample_method_grid |>
  tidyr::crossing(taxon_universe) |>
  dplyr::left_join(expected_truth, by = c("taxonomic_rank", "taxon")) |>
  dplyr::mutate(
    expected_abundance = dplyr::case_when(
      matrix_type %in% c("mock_only", "mock_plus_stool") &
        !is.na(expected_abundance) ~ expected_abundance,
      TRUE ~ 0
    )
  ) |>
  dplyr::left_join(
    taxa_abundance,
    by = c(
      "run_id",
      "sample_id",
      "DNA_kit",
      "replicate",
      "matrix_type",
      "pipeline",
      "preprocess_method",
      "read_strategy",
      "taxonomy_method",
      "BLAST",
      "Paired",
      "Trimm",
      "taxonomic_rank",
      "taxon"
    )
  ) |>
  dplyr::mutate(
    count = tidyr::replace_na(count, 0),
    total_reads = tidyr::replace_na(total_reads, 0),
    relative_abundance = tidyr::replace_na(relative_abundance, 0)
  ) |>
  dplyr::select(
    run_id,
    sample_id,
    DNA_kit,
    replicate,
    matrix_type,
    pipeline,
    preprocess_method,
    read_strategy,
    taxonomy_method,
    BLAST,
    Paired,
    Trimm,
    taxonomic_rank,
    taxon,
    count,
    total_reads,
    relative_abundance,
    expected_abundance
  )

benchmark_long_species <- benchmark_long |>
  dplyr::filter(taxonomic_rank == "Species")

benchmark_long_genus <- benchmark_long |>
  dplyr::filter(taxonomic_rank == "Genus")

# readr::write_tsv(benchmark_long,
#                  file.path(table_dir, "benchmark_long_genus_species.tsv"))
# readr::write_tsv(benchmark_long_species,
#                  file.path(table_dir, "benchmark_long_species.tsv"))
# readr::write_tsv(benchmark_long_genus,
#                  file.path(table_dir, "benchmark_long_genus.tsv"))
```

# 7. Figure4A: Species-level taxa plot
## Original Plot 

```{R}


# Mock target genus 목록 생성
mock_genera <- unique(stringr::str_replace(expected_species, "_.*$", ""))

# Species-level table에서 genus 추출
benchmark_long_species <- benchmark_long_species |>
  dplyr::mutate(
    Genus = stringr::str_extract(taxon, "^[A-Za-z]+")
  )

# 실제 데이터에서 관찰된 mock genus 소속 taxa만 추출
observed_mock_taxa <- benchmark_long_species |>
  dplyr::filter(
    Genus %in% mock_genera,
    taxon != "Unclassified",
    relative_abundance > 0
  ) |>
  dplyr::group_by(Genus, taxon) |>
  dplyr::summarise(
    total_abundance = sum(relative_abundance, na.rm = TRUE),
    max_abundance = max(relative_abundance, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::filter(max_abundance >= 0.0001) |>
  dplyr::arrange(match(Genus, mock_genera), dplyr::desc(total_abundance), taxon)

# Genus-level로만 잡힌 경우 표시할 Genus_g label 자동 생성
observed_genus_labels <- observed_mock_taxa |>
  dplyr::distinct(Genus) |>
  dplyr::mutate(taxon = paste0(Genus, "_g")) |>
  dplyr::select(Genus, taxon)

# 실제로 plot에 표시할 taxa level 자동 생성
sp_level <- dplyr::bind_rows(
  observed_mock_taxa |> dplyr::select(Genus, taxon),
  observed_genus_labels
) |>
  dplyr::distinct(Genus, taxon) |>
  dplyr::arrange(match(Genus, mock_genera), taxon) |>
  dplyr::pull(taxon)

# Genus별 기본 색상 지정
genus_base_color <- c(
  Aspergillus = "#E41A1CFF",
  Candida = "#377EB8FF",
  Cryptococcus = "#4DAF4AFF",
  Cutaneotrichosporon = "#984EA3FF",
  Fusarium = "#FF7F00FF",
  Malassezia = "#FFFF33FF",
  Nakaseomyces = "#A65628FF",
  Penicillium = "#f55cacFF",
  Saccharomyces = "#E6AB02FF",
  Trichophyton = "#1B9E77FF"
)

# 관찰된 taxa 수에 따라 genus 색상 계열 내에서 자동 tint 생성
taxon_color_df <- tibble::tibble(taxon = sp_level) |>
  dplyr::mutate(
    Genus = stringr::str_extract(taxon, "^[A-Za-z]+"),
    base_color = genus_base_color[Genus]
  ) |>
  dplyr::group_by(Genus) |>
  dplyr::arrange(taxon, .by_group = TRUE) |>
  dplyr::mutate(
    rank = dplyr::row_number(),
    n_in_genus = dplyr::n(),
    tint_amount = dplyr::if_else(n_in_genus > 1, (rank - 1) / (n_in_genus - 1) * 0.65, 0)  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    color = purrr::map2_chr(
      base_color,
      tint_amount,
      ~ ifelse(.y == 0, .x, colorspace::lighten(.x, amount = .y))
    )
  )

taxa_colors <- stats::setNames(taxon_color_df$color, taxon_color_df$taxon)
taxa_colors <- c(taxa_colors, Other = "#999999FF")
```



```{r}
# 실제 데이터 기반으로 표시 taxa 결정
plot_df <- benchmark_long_species |>
  dplyr::mutate(
    Species2 = dplyr::case_when(
      taxon %in% sp_level ~ taxon,
      Genus %in% mock_genera ~ paste0(Genus, "_g"),
      TRUE ~ "Other"
    ),
    Species2 = dplyr::if_else(Species2 %in% sp_level, Species2, "Other"),
    Species2 = factor(Species2, levels = c(sp_level, "Other")),
    Xaxis = paste(run_id, sample_id, sep = "_")
  ) |>
  dplyr::group_by(
    run_id, sample_id, Xaxis, DNA_kit, replicate, matrix_type,
    taxonomy_method, BLAST, Trimm, Paired, Species2
  ) |>
  dplyr::summarise(
    relative_abundance = sum(relative_abundance, na.rm = TRUE),
    .groups = "drop"
  )

readr::write_tsv(observed_mock_taxa, file.path(table_dir, "figure1_observed_mock_taxa.tsv"))
readr::write_tsv(taxon_color_df, file.path(table_dir, "figure1_taxa_color_table.tsv"))
readr::write_tsv(plot_df, file.path(table_dir, "figure1_species_taxa_plot_data.tsv"))
```



```{r}
# Mon Jul 27 10:32:38 2026 ------------------------------
# 
# p_taxa <- ggplot2::ggplot(
#   plot_df,
#   ggplot2::aes(x = sample_id, y = relative_abundance, fill = Species2)
# ) +
#   ggplot2::geom_col(width = 0.95) +
#   ggplot2::scale_fill_manual(values = taxa_colors, drop = FALSE) +
#   ggplot2::scale_y_continuous(expand = c(0, 0)) +
#   ggh4x::facet_nested(
#    ~ taxonomy_method + Trimm + Paired  ,
#   scales = "free_x",
#   space = "free_x"
# ) +
#   ggplot2::labs(x = NULL, y = "Relative abundance", fill = "Taxon") +
#   ggplot2::theme_classic(base_size = 12) +
#   ggplot2::theme(
#     axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5, color = "black", size = 7),
#     axis.text.y = ggplot2::element_text(color = "black"),
#     axis.title.y = ggplot2::element_text(color = "black", face = "bold"),
#     axis.line = ggplot2::element_line(color = "black", linewidth = 0.5),
#     axis.ticks = ggplot2::element_line(color = "black", linewidth = 0.5),
#     strip.background = ggplot2::element_blank(),
#     strip.text.x = ggplot2::element_text(face = "bold", color = "black"),
#     legend.position = "right",
#     legend.key.size = grid::unit(0.35, "cm")
#   )

p_taxa <- ggplot2::ggplot(
  plot_df,
  ggplot2::aes(x = sample_id, y = relative_abundance, fill = Species2)
) +
  ggplot2::geom_col(width = 0.95) +
  ggplot2::scale_fill_manual(values = taxa_colors, drop = FALSE) +
  ggplot2::scale_y_continuous(expand = c(0, 0)) +
  ggh4x::facet_nested(
    ~ taxonomy_method + Trimm + Paired,
    scales = "free_x",
    space = "free_x"
  ) +
  ggplot2::labs(x = NULL, y = "Relative abundance", fill = "Taxon") +
  ggplot2::guides(
    fill = ggplot2::guide_legend(
      nrow = 3,
      byrow = TRUE
    )
  ) +
  ggplot2::theme_classic(base_size = 12) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5, color = "black", size = 7),
    axis.text.y = ggplot2::element_text(color = "black"),
    axis.title.y = ggplot2::element_text(color = "black", face = "bold"),
    axis.line = ggplot2::element_line(color = "black", linewidth = 0.5),
    axis.ticks = ggplot2::element_line(color = "black", linewidth = 0.5),
    strip.background = ggplot2::element_blank(),
    strip.text.x = ggplot2::element_text(face = "bold", color = "black"),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.key.size = grid::unit(0.35, "cm"),
    legend.text = ggplot2::element_text(size = 7),
    legend.title = ggplot2::element_text(size = 9, face = "bold")
  )



p_taxa
```



```{r}
plot_df2 <- plot_df %>% 
  mutate(method = paste0(Trimm,"_", Paired))


p_taxa2 <- ggplot2::ggplot(
  plot_df2,
  ggplot2::aes(x = method, y = relative_abundance, fill = Species2)
) +
  ggplot2::geom_col(width = 0.95, position = "fill") +
  ggplot2::scale_fill_manual(values = taxa_colors, drop = FALSE) +
  ggplot2::scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = c(0, 0)
  ) +
  ggh4x::facet_nested(
    ~ taxonomy_method,
    scales = "free_x",
    space = "free_x"
  ) +
  ggplot2::labs(x = NULL, y = "Relative abundance (%)", fill = "Taxon") +
  ggplot2::theme_classic(base_size = 12) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5, color = "black", size = 7),
    axis.text.y = ggplot2::element_text(color = "black"),
    axis.title.y = ggplot2::element_text(color = "black", face = "bold"),
    axis.line = ggplot2::element_line(color = "black", linewidth = 0.5),
    axis.ticks = ggplot2::element_line(color = "black", linewidth = 0.5),
    strip.background = ggplot2::element_blank(),
    strip.text.x = ggplot2::element_text(face = "bold", color = "black"),
    legend.position = "right",
    legend.key.size = grid::unit(0.35, "cm")
  )

p_taxa2
```


## Ground truth 값 추가 후 확인 
```{r}

# Ground truth를 taxa plot용 데이터 구조로 변환

colnames(plot_df)

ground_truth_plot_df <- expected_truth |>
  dplyr::filter(taxonomic_rank == "Species") |>
  dplyr::mutate(
    Genus = stringr::str_extract(taxon, "^[A-Za-z]+"),
    Species2 = dplyr::case_when(
      taxon %in% sp_level ~ taxon,
      Genus %in% mock_genera ~ paste0(Genus, "_g"),
      TRUE ~ "Other"
    ),
    Species2 = dplyr::if_else(Species2 %in% sp_level, Species2, "Other"),
    run_id = "Ground truth",
    sample_id = "Ground truth",
    Xaxis = "Ground truth",
    DNA_kit = "Ground truth",
    replicate = 1,
    matrix_type = "Ground truth",
    taxonomy_method = "Ground truth",
    BLAST = "Ground truth",
    Trimm = "Ground truth",
    Paired = "Ground truth",
    relative_abundance = expected_abundance
  ) |>
  dplyr::group_by(
    run_id, sample_id, Xaxis, DNA_kit, replicate, matrix_type,
    taxonomy_method, BLAST, Trimm, Paired, Species2
  ) |>
  dplyr::summarise(
    relative_abundance = sum(relative_abundance, na.rm = TRUE),
    .groups = "drop"
  )

# 기존 plot_df 앞에 Ground truth 추가
plot_df_with_truth <- dplyr::bind_rows(
  ground_truth_plot_df,
  plot_df
) |>
  dplyr::mutate(
    Species2 = factor(Species2, levels = c(sp_level, "Other")),
    sample_id = factor(
      sample_id,
      levels = c(
        "Ground truth",
        plot_df |>
          dplyr::distinct(sample_id) |>
          dplyr::pull(sample_id)
      )
    ),
    taxonomy_method = factor(
      taxonomy_method,
      levels = c(
        "Ground truth",
        setdiff(unique(plot_df$taxonomy_method), "Ground truth")
      )
    ),
    Trimm = factor(
      Trimm,
      levels = c(
        "Ground truth",
        setdiff(unique(plot_df$Trimm), "Ground truth")
      )
    ),
    Paired = factor(
      Paired,
      levels = c(
        "Ground truth",
        setdiff(unique(plot_df$Paired), "Ground truth")
      )
    )
  )

readr::write_tsv(
  plot_df_with_truth,
  file.path(table_dir, "figure1_species_taxa_plot_data_with_ground_truth.tsv")
)

```


```{r, fig.width=12}
# Mon Jul 27 10:32:38 2026 ------------------------------

p_taxa_gt <- ggplot2::ggplot(
  plot_df_with_truth,
  ggplot2::aes(x = sample_id, y = relative_abundance, fill = Species2)
) +
  ggplot2::geom_col(width = 0.95) +
  ggplot2::scale_fill_manual(values = taxa_colors, drop = FALSE) +
  ggplot2::scale_y_continuous(expand = c(0, 0)) +
  ggh4x::facet_nested(
   ~ taxonomy_method + Trimm + Paired  ,
  scales = "free_x",
  space = "free_x"
) +
  ggplot2::labs(x = NULL, y = "Relative abundance", fill = "Taxon") +
  ggplot2::theme_classic(base_size = 12) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5, color = "black", size = 7),
    axis.text.y = ggplot2::element_text(color = "black"),
    axis.title.y = ggplot2::element_text(color = "black", face = "bold"),
    axis.line = ggplot2::element_line(color = "black", linewidth = 0.5),
    axis.ticks = ggplot2::element_line(color = "black", linewidth = 0.5),
    strip.background = ggplot2::element_blank(),
    strip.text.x = ggplot2::element_text(face = "bold", color = "black"),
    legend.position = "right",
    legend.key.size = grid::unit(0.35, "cm")
  )


p_taxa_gt
```



```{r}
plot_df_with_truth2 <- plot_df_with_truth %>% 
  dplyr::mutate(
    method = dplyr::if_else(
      taxonomy_method == "Ground truth",
      "Ground truth",
      paste0(Trimm, "_", Paired)
    )
  )

p_taxa2_gt <- ggplot2::ggplot(
  plot_df_with_truth2,
  ggplot2::aes(x = method, y = relative_abundance, fill = Species2)
) +
  ggplot2::geom_col(width = 0.95, position = "fill") +
  ggplot2::scale_fill_manual(values = taxa_colors, drop = FALSE) +
  ggplot2::scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = c(0, 0)
  ) +
  ggh4x::facet_nested(
    ~ taxonomy_method,
    scales = "free_x",
    space = "free_x"
  ) +
  ggplot2::labs(x = NULL, y = "Relative abundance (%)", fill = "Taxon") +
  ggplot2::theme_classic(base_size = 12) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5, color = "black", size = 7),
    axis.text.y = ggplot2::element_text(color = "black"),
    axis.title.y = ggplot2::element_text(color = "black", face = "bold"),
    axis.line = ggplot2::element_line(color = "black", linewidth = 0.5),
    axis.ticks = ggplot2::element_line(color = "black", linewidth = 0.5),
    strip.background = ggplot2::element_blank(),
    strip.text.x = ggplot2::element_text(face = "bold", color = "black"),
    legend.position = "right",
    legend.key.size = grid::unit(0.35, "cm")
  )

p_taxa2_gt
```




# 8. Figure4B: Ground-truth accuracy metrics

```{r}
# Sat Jul 25 17:38:44 2026 ------------------------------
# Mon Jul 27 10:45:57 2026 ------------------------------


method_colors <- c(
  "ITSx_PE"     = "#66C2A5",
  "ITSx_SE"     = "#FC8D62",
  "Cutadapt_PE" = "#8DA0CB",
  "Cutadapt_SE" = "#E78AC3"
)
# 샘플링
matrix_keep <- "mock_only"

metric_levels_keep <- c(
  "F1-score",
  "R²",
  "Bray–Curtis distance"
)


```

```{r metric-calc-plot, fig.width=16, fig.height=10}
# Ground-truth 기반 정확도 지표를 계산하고 저장
metrics_main <- calc_metrics(benchmark_long, threshold_value = 0, threshold_label = "count_gt_0")
metrics_sensitivity <- calc_metrics(benchmark_long, threshold_value = 0, threshold_label = "relative_abundance_ge_0")

readr::write_tsv(metrics_main, file.path(table_dir, "benchmark_metrics_main_count_gt_0.tsv"))
readr::write_tsv(metrics_sensitivity, file.path(table_dir, "benchmark_metrics_sensitivity_ra_ge_0.tsv"))

# Species level 지표만 선택하고 분석 조건 순서를 지정
meta2 <- metrics_main |>
  dplyr::filter(taxonomic_rank == "Species") |>
  dplyr::filter(taxonomy_method != "Mock type-strain BLAST") |>
  dplyr::mutate(
    DNA_kit = factor(DNA_kit, levels = c("DNeasy", "QIAamp", "Zymo")),
    BLAST = factor(BLAST, levels = c("No_BLAST", "BLAST")),
    taxonomy_method = factor(taxonomy_method, levels = c("Mock type-strain BLAST", "UNITE", "UNITE+RefSeq BLAST")),
    Trimm = factor(Trimm, levels = c("ITSxpress", "Cutadapt")),
    Paired = factor(Paired, levels = c("PE", "SE")),
    Method = paste(Trimm, Paired, sep = "_"),
    Method = dplyr::recode(
      Method,
      "ITSxpress_PE" = "ITSx_PE",
      "ITSxpress_SE" = "ITSx_SE",
      "Cutadapt_PE" = "Cutadapt_PE",
      "Cutadapt_SE" = "Cutadapt_SE"
    ),
    Method = factor(Method, levels = c("ITSx_PE", "ITSx_SE", "Cutadapt_PE", "Cutadapt_SE"))
  )

# Plot용 long-format table 생성
metric_plot_df <- meta2 |>
  dplyr::select(
    run_id, sample_id, DNA_kit, replicate, matrix_type,
    Trimm, Paired, Method, taxonomy_method, BLAST,
    F1, R2, Bray_Curtis, precision, recall, false_discovery_rate
  ) |>
  tidyr::pivot_longer(
    cols = c(F1, R2, Bray_Curtis, precision, recall, false_discovery_rate),
    names_to = "metric",
    values_to = "value"
  ) |>
  dplyr::mutate(
    metric = dplyr::recode(
      metric,
      "F1" = "F1-score",
      "R2" = "R²",
      "Bray_Curtis" = "Bray–Curtis",
      "precision" = "Precision",
      "recall" = "Recall",
      "false_discovery_rate" = "FDR"
    ),
    metric = factor(
      metric,
      levels = c("F1-score", "R²", "Bray–Curtis", "Precision", "Recall", "FDR")
    )
  )

# readr::write_tsv(meta2, file.path(table_dir, "figure2_species_metric_table.tsv"))
# readr::write_tsv(metric_plot_df, file.path(table_dir, "figure2_species_metric_plot_data.tsv"))
```


```{r, fig.width=16}
# 조건별 평균과 표준오차를 계산
metric_summary_df <- metric_plot_df |>
  dplyr::group_by(metric, taxonomy_method, Method) |>
  dplyr::summarise(
    mean_value = mean(value, na.rm = TRUE),
    sd_value = stats::sd(value, na.rm = TRUE),
    n = sum(!is.na(value)),
    se_value = sd_value / sqrt(n),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    metric = factor(
      metric,
      levels = c("F1-score", "R²", "Bray–Curtis", "Precision", "Recall", "FDR")
    ),
    taxonomy_method = factor(taxonomy_method, levels = c("Mock type-strain BLAST", "UNITE", "UNITE+RefSeq BLAST")),
    Method = factor(Method, levels = c("ITSx_PE", "ITSx_SE", "Cutadapt_PE", "Cutadapt_SE"))
  )

readr::write_tsv(metric_summary_df, file.path(table_dir, "figure2_species_metric_summary_barplot_0.tsv"))

# 분석 조건별 색상 지정
method_colors <- c(
  "ITSx_PE" = "#66C2A5",
  "ITSx_SE" = "#FC8D62",
  "Cutadapt_PE" = "#8DA0CB",
  "Cutadapt_SE" = "#E78AC3"
)

# Figure B 스타일의 metric summary barplot 생성
p_metrics <- ggplot2::ggplot(
  metric_summary_df,
  ggplot2::aes(x = Method, y = mean_value, fill = Method)
) +
  ggplot2::geom_col(width = 0.65, color = NA) +
  ggplot2::geom_errorbar(
    ggplot2::aes(ymin = mean_value - se_value, ymax = mean_value + se_value),
    width = 0.18,
    linewidth = 0.4,
    color = "black"
  ) +
  ggplot2::scale_fill_manual(values = method_colors) +
 ggh4x::facet_wrap2(
    ggplot2::vars(metric, taxonomy_method),
    nrow = 2,
    scales = "free_y",
    strip = ggh4x::strip_nested()
  ) +
  ggplot2::labs(x = NULL, y = NULL) +
  ggplot2::theme_classic(base_size = 12) +
  ggplot2::theme(
    legend.position = "none",
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, vjust = 1, color = "black"),
    axis.text.y = ggplot2::element_text(color = "black"),
    axis.line = ggplot2::element_line(color = "black", linewidth = 0.5),
    axis.ticks = ggplot2::element_line(color = "black", linewidth = 0.5),
    strip.background = ggplot2::element_blank(),
    strip.text.x = ggplot2::element_text(face = "plain", color = "black", size = 11),
    panel.spacing.x = grid::unit(1.1, "lines")
  )

p_metrics

ggplot2::ggsave(file.path(figure_dir, "figure2_ground_truth_accuracy_metrics_barplot.png"), p_metrics, width = 26, height = 5, dpi = 300)
# ggplot2::ggsave(file.path(figure_dir, "figure2_ground_truth_accuracy_metrics_barplot.pdf"), p_metrics, width = 26, height = 5)
```

# 9. Figure4C: Random forest importance and PERMANOVA R²

## 1) function
```{r}

meta2[is.na(meta2$BLAST), "BLAST"] <- "BLAST"

plot_effect <- function(df) {
  ggplot(df, aes(
    x = stats::reorder(.data$Variable, .data$IncMSE),
    y = .data$IncMSE
  )) +
    ggplot2::geom_point(fill = "steelblue") +
    ggplot2::coord_flip() +
    fig_theme(base_size = 13)
}

```

## 2) ML
```{r , fig.width=11, fig.height=7}

## Machine learning 기반 확인 
rf_r2_out <- make_rf_importance(meta2, "R2")
rf_f1_out <- make_rf_importance(meta2, "F1")


rf_r2 <- rf_r2_out$model
rf_f1 <- rf_f1_out$model

importance_r2.1 <- rf_r2_out$importance
importance_f1.1 <- rf_f1_out$importance


importance_all <- dplyr::bind_rows(
  importance_r2.1 |> dplyr::mutate(outcome = "R²"),
  importance_f1.1 |> dplyr::mutate(outcome = "F1-score")
)
readr::write_tsv(importance_all, file.path(table_dir, "rf_importance_combined.tsv"))
```

## 3) PERMANOVA
```{r}
community_wide <- benchmark_long_species |>
  dplyr::mutate(sample_method_id = paste(run_id, sample_id, taxonomy_method, sep = "__")) |>
  dplyr::select(sample_method_id, taxon, relative_abundance) |>
  dplyr::group_by(sample_method_id, taxon) |>
  dplyr::summarise(relative_abundance = sum(relative_abundance, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = taxon, values_from = relative_abundance, values_fill = 0)

perma_meta <- benchmark_long_species |>
  dplyr::mutate(sample_method_id = paste(run_id, sample_id, taxonomy_method, sep = "__")) |>
  dplyr::distinct(sample_method_id, run_id, sample_id, DNA_kit, BLAST, Paired, Trimm, matrix_type)

community_mat <- community_wide |>
  dplyr::arrange(sample_method_id) |>
  tibble::column_to_rownames("sample_method_id") |>
  as.matrix()

perma_meta <- perma_meta |>
  dplyr::filter(sample_method_id %in% rownames(community_mat)) |>
  dplyr::arrange(match(sample_method_id, rownames(community_mat)))

perma_vars <- c("DNA_kit", "BLAST", "Paired", "Trimm")
perma_vars <- perma_vars[purrr::map_lgl(perma_vars, ~ dplyr::n_distinct(perma_meta[[.x]], na.rm = TRUE) > 1)]



## PERMANOVA
set.seed(123)
dist_matrix <- vegan::vegdist(community_mat, method = "bray")
per_res <- vegan::adonis2(
  formula = stats::as.formula(paste("dist_matrix ~", paste(perma_vars, collapse = " + "))),
  data = perma_meta,
  permutations = 9999,
  by = "margin", 
  strata = perma_meta$sample_id
)

## 결과 정리 
perma_out <- per_res |>
  as.data.frame() |>
  tibble::rownames_to_column("Variable") |>
  dplyr::filter(!Variable %in% c("Residual", "Total")) |>
  dplyr::rename(p_value = `Pr(>F)`)

readr::write_tsv(perma_out, file.path(table_dir, "permanova_R2.tsv"))



```


## 4) Figure

- Permutation importance  = %IncMSE
```{r, fig.width=12, fig.height=3}
p_rf_r2 <- plot_effect(importance_r2.1) +  labs(x = NULL, y = "%IncMSE", title = "R²")

p_rf_f1 <- plot_effect(importance_f1.1) +  labs(x = NULL, y = "%IncMSE", title = "F1-score")

p_permanova <- ggplot(perma_out, aes( x = stats::reorder(Variable, R2), y = R2)) +
  ggplot2::geom_point(fill = "steelblue") + ggplot2::coord_flip() +
  fig_theme(base_size = 13) + ggplot2::labs(x = NULL, y = "Explained variance (R²)", title = "PERMANOVA")



p_fig4c <- p_rf_r2 | p_rf_f1 | p_permanova

p_fig4c

```



# 10. Combined figure

```{r combined-figure, fig.width=12, fig.height=12}
fig_combined <- p_taxa /
  p_metrics /
  p_fig4c +
  patchwork::plot_layout(heights = c(1.2, 1.4, 0.7)) 

fig_combined

ggplot2::ggsave(file.path(figure_dir, "figure_combined_ITSxpress_PE_corrected.png"), fig_combined, width = 12, height = 12, 
                dpi = 300)
ggplot2::ggsave(file.path(figure_dir, "figure_combined_ITSxpress_PE_corrected.pdf"), fig_combined, width = 12, height = 12)
```



# 11. Session information

```{r session-info}
sessionInfo()
```




