---
title: "Figure 4 pilot: ITSxpress PE taxonomy benchmark"
author: "Kim So-Yeon"
date: "2026-07-24"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)
options(stringsAsFactors = FALSE)
```

# 1. Brief analysis plan

This Rmd processes one completed ITSdetector run and compares two taxonomy versions using the same ASV abundance table.

- UNITE before BLAST: `taxonomy_normalized.tsv`
- UNITE+BLAST after refinement: `taxonomy_blast.tsv` with normalized rank columns
- Species names are harmonized before comparison with the expected mock composition.
- Main detection rule: `count > 0`
- Sensitivity detection rule: `relative_abundance >= 0.001`
- PBS negative controls are retained in composition plots but excluded from ground-truth accuracy summaries.

# 2. Load packages and paths

```{r load-packages-paths}
required_pkgs <- c(
  "tidyverse", "qiime2R", "ggplot2", "ggh4x", "vegan",
  "randomForest", "patchwork", "scales", "rmarkdown"
)

missing_pkgs <- required_pkgs[!purrr::map_lgl(required_pkgs, requireNamespace, quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  stop("Missing packages: ", paste(missing_pkgs, collapse = ", "))
}

project_dir <- "/data/home2/ksy/260724_GutMyco_Fungal_Mock_Paper"
rproj_dir <- file.path(project_dir, "Rproj")
run_id <- "fig4_q20_PE_itsxpress_unite"
run_dir <- file.path(project_dir, "Output", run_id)

analysis_dir <- file.path(rproj_dir, "Output", "fig4_pilot_itsxpress_PE_corrected")
table_dir <- file.path(analysis_dir, "tables")
figure_dir <- file.path(analysis_dir, "figures")

dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

paths <- list(
  table_qza = file.path(run_dir, "dada2", "itsxpress", "paired", "table.qza"),
  repseq_qza = file.path(run_dir, "dada2", "itsxpress", "paired", "repseq.qza"),
  denoising_stats_qza = file.path(run_dir, "dada2", "itsxpress", "paired", "denoising_stats.qza"),
  taxonomy_unite_normalized = file.path(run_dir, "taxonomy_blast", "itsxpress", "paired", "unite_ver10_99", "taxonomy_normalized.tsv"),
  taxonomy_blast_final = file.path(run_dir, "taxonomy_blast", "itsxpress", "paired", "unite_ver10_99", "taxonomy_blast.tsv"),
  taxonomy_blast_qiime = file.path(run_dir, "taxonomy_blast", "itsxpress", "paired", "unite_ver10_99", "taxonomy_blast_qiime.tsv"),
  blast_evidence = file.path(run_dir, "taxonomy_blast", "itsxpress", "paired", "unite_ver10_99", "taxonomy_blast_evidence.tsv"),
  blast_report = file.path(run_dir, "taxonomy_blast", "itsxpress", "paired", "unite_ver10_99", "taxonomy_blast_report.tsv"),
  sample_metadata = file.path(project_dir, "Input", "metadata_mock_stool_and_mock_260724.tsv"),
  cleanup_qc_summary = file.path(run_dir, "cleanup_qc", "cleanup_qc_summary.tsv")
)

path_check <- tibble::tibble(
  file_key = names(paths),
  path = unlist(paths),
  exists = file.exists(unlist(paths))
)

readr::write_tsv(path_check, file.path(table_dir, "input_file_check.tsv"))
print(path_check)

missing_required <- path_check |>
  dplyr::filter(
    .data$file_key %in% c("table_qza", "taxonomy_unite_normalized", "taxonomy_blast_final", "sample_metadata"),
    !.data$exists
  )

if (nrow(missing_required) > 0) {
  stop("Missing required input files:\n", paste(missing_required$path, collapse = "\n"))
}
```

# 3. Define expected mock composition and plotting style

```{r expected-style}
expected_species <- c(
  "Aspergillus_fumigatus",
  "Cryptococcus_neoformans",
  "Trichophyton_interdigitale",
  "Penicillium_chrysogenum",
  "Fusarium_keratoplasticum",
  "Candida_albicans",
  "Nakaseomyces_glabratus",
  "Malassezia_globosa",
  "Saccharomyces_cerevisiae",
  "Cutaneotrichosporon_dermatis"
)

expected_genus <- c(
  "Aspergillus", "Cryptococcus", "Trichophyton", "Penicillium", "Fusarium",
  "Candida", "Nakaseomyces", "Malassezia", "Saccharomyces", "Cutaneotrichosporon"
)

expected_truth <- dplyr::bind_rows(
  tibble::tibble(taxonomic_rank = "Species", taxon = expected_species, expected_abundance = 0.1),
  tibble::tibble(taxonomic_rank = "Genus", taxon = expected_genus, expected_abundance = 0.1)
)

nomenclature_map <- tibble::tribble(
  ~rank, ~input_name, ~harmonized_name, ~note,
  "Species", "Candida_glabrata", "Nakaseomyces_glabratus", "Current name used for comparison.",
  "Species", "Nakaseomyces_glabrata", "Nakaseomyces_glabratus", "Spelling harmonization.",
  "Species", "Cryptococcus_neoformans_var_grubii", "Cryptococcus_neoformans", "Variety collapsed to species level.",
  "Species", "Cryptococcus_neoformans_var._grubii", "Cryptococcus_neoformans", "Variety collapsed to species level.",
  "Species", "Penicillium_chrysogenum_species_complex", "Penicillium_chrysogenum", "Species complex collapsed to mock target name."
)

readr::write_tsv(expected_truth, file.path(table_dir, "expected_mock_composition.tsv"))
readr::write_tsv(nomenclature_map, file.path(table_dir, "taxonomy_name_harmonization_map.tsv"))

tax_col <- c(
  "#E41A1CFF", "#377EB8FF", "#4DAF4AFF", "#984EA3FF", "#FF7F00FF",
  "#FFFF33FF", "#A65628FF", "#f55cacFF", "#E6AB02FF", "#1B9E77FF", "#999999FF"
)

met_col <- c(
  "#66C2A5", "#FC8D62", "#8DA0CB", "#E78AC3",
  "#A6D854", "#FFD92F", "#E5C494", "#B3B3B3"
)

fig_theme <- function(base_size = 12) {
  ggplot2::theme_classic(base_size = base_size) +
    ggplot2::theme(
      panel.background = ggplot2::element_blank(),
      plot.background = ggplot2::element_blank(),
      panel.grid = ggplot2::element_blank(),
      axis.line = ggplot2::element_line(color = "black", linewidth = 0.5),
      axis.ticks = ggplot2::element_line(color = "black", linewidth = 0.5),
      axis.text.x.bottom = ggplot2::element_text(angle = 45, hjust = 1.1, vjust = 0.9),
      legend.title = ggplot2::element_blank(),
      strip.background = ggplot2::element_blank(),
      strip.text = ggplot2::element_text(face = "bold")
    )
}
```

# 4. Helper functions

```{r helper-functions}
clean_taxon_value <- function(x) {
  x |>
    as.character() |>
    stringr::str_replace_all("^.__", "") |>
    stringr::str_replace_all(" ", "_") |>
    stringr::str_replace_all(";.*$", "") |>
    stringr::str_replace_all("^_+|_+$", "") |>
    dplyr::na_if("")
}

harmonize_taxon <- function(x, rank = "Species") {
  x1 <- clean_taxon_value(x)

  dplyr::case_when(
    rank == "Species" & x1 %in% c("Candida_glabrata", "Nakaseomyces_glabrata") ~ "Nakaseomyces_glabratus",
    rank == "Species" & x1 %in% c("Cryptococcus_neoformans_var_grubii", "Cryptococcus_neoformans_var._grubii") ~ "Cryptococcus_neoformans",
    rank == "Species" & x1 == "Penicillium_chrysogenum_species_complex" ~ "Penicillium_chrysogenum",
    TRUE ~ x1
  )
}

parse_qiime_taxon <- function(taxon_string, rank_index) {
  split_mat <- stringr::str_split_fixed(taxon_string, ";", 7)
  clean_taxon_value(split_mat[, rank_index])
}

read_taxonomy_table <- function(path, taxonomy_method_label) {
  tax <- readr::read_tsv(path, show_col_types = FALSE)

  if ("Feature ID" %in% colnames(tax)) {
    tax <- dplyr::rename(tax, feature_id = `Feature ID`)
  }

  if ("ASV" %in% colnames(tax)) {
    tax <- dplyr::rename(tax, feature_id = ASV)
  }

  if (!"feature_id" %in% colnames(tax)) {
    stop("Taxonomy table has no Feature ID or ASV column: ", path)
  }

  if (!"Genus" %in% colnames(tax) && "Taxon" %in% colnames(tax)) {
    tax <- dplyr::mutate(tax, Genus = parse_qiime_taxon(.data$Taxon, 6))
  }

  if (!"Species" %in% colnames(tax) && "Taxon" %in% colnames(tax)) {
    tax <- dplyr::mutate(tax, Species = parse_qiime_taxon(.data$Taxon, 7))
  }

  if (!all(c("Genus", "Species") %in% colnames(tax))) {
    stop("Taxonomy table must contain Genus and Species columns or a parsable Taxon column: ", path)
  }

  if (!"Taxon" %in% colnames(tax)) {
    tax <- dplyr::mutate(tax, Taxon = NA_character_)
  }

  tax |>
    dplyr::transmute(
      feature_id = as.character(.data$feature_id),
      taxonomy_method = taxonomy_method_label,
      Genus = harmonize_taxon(.data$Genus, rank = "Genus"),
      Species = harmonize_taxon(.data$Species, rank = "Species"),
      taxon_string = as.character(.data$Taxon)
    ) |>
    dplyr::mutate(
      Genus = dplyr::if_else(is.na(.data$Genus) | stringr::str_detect(.data$Genus, "_g$"), "Unclassified", .data$Genus),
      Species = dplyr::if_else(is.na(.data$Species) | stringr::str_detect(.data$Species, "_g$|_s$"), "Unclassified", .data$Species)
    )
}

standardize_metadata <- function(metadata_df) {
  meta <- metadata_df

  if ("sample" %in% colnames(meta) && !"sample_id" %in% colnames(meta)) {
    meta <- dplyr::rename(meta, sample_id = sample)
  }

  if ("kit_code" %in% colnames(meta) && !"kit" %in% colnames(meta)) {
    meta <- dplyr::rename(meta, kit = kit_code)
  }

  meta |>
    dplyr::mutate(
      sample_id = as.character(.data$sample_id),
      sample_id = stringr::str_replace(.data$sample_id, "^X", ""),
      DNA_kit = dplyr::recode(
        as.character(.data$kit),
        "D" = "DNeasy",
        "Q" = "QIAamp",
        "Z" = "Zymo",
        .default = as.character(.data$kit)
      ),
      replicate = as.character(.data$replicate),
      matrix_type = as.character(.data$matrix_type)
    ) |>
    dplyr::select(.data$sample_id, .data$DNA_kit, .data$replicate, .data$matrix_type, dplyr::everything()) |>
    dplyr::distinct(.data$sample_id, .keep_all = TRUE)
}

feature_table_to_long <- function(feature_table_obj, taxonomy_feature_ids) {
  ft <- as.data.frame(feature_table_obj)
  row_match <- sum(rownames(ft) %in% taxonomy_feature_ids)
  col_match <- sum(colnames(ft) %in% taxonomy_feature_ids)

  if (col_match > row_match) {
    ft <- t(ft) |> as.data.frame()
  }

  ft |>
    tibble::rownames_to_column("feature_id") |>
    tidyr::pivot_longer(
      cols = - .data$feature_id,
      names_to = "sample_id",
      values_to = "count"
    ) |>
    dplyr::mutate(
      sample_id = stringr::str_replace(as.character(.data$sample_id), "^X", ""),
      count = as.numeric(.data$count)
    )
}

calc_bray <- function(observed, expected) {
  if (sum(observed, na.rm = TRUE) + sum(expected, na.rm = TRUE) == 0) {
    return(NA_real_)
  }
  as.numeric(vegan::vegdist(rbind(observed, expected), method = "bray")[1])
}

calc_r2 <- function(observed, expected) {
  if (stats::sd(observed, na.rm = TRUE) == 0 || stats::sd(expected, na.rm = TRUE) == 0) {
    return(NA_real_)
  }
  stats::cor(observed, expected, use = "complete.obs")^2
}

calc_metrics <- function(df, threshold_value, threshold_label) {
  df |>
    dplyr::mutate(
      detected = .data$relative_abundance > threshold_value,
      expected_detected = .data$expected_abundance > 0
    ) |>
    dplyr::filter(.data$matrix_type %in% c("mock_only", "mock_plus_stool")) |>
    dplyr::group_by(
      .data$sample_id, .data$DNA_kit, .data$replicate, .data$matrix_type,
      .data$pipeline, .data$preprocess_method, .data$read_strategy,
      .data$taxonomy_method, .data$BLAST, .data$Paired, .data$Trimm, .data$taxonomic_rank
    ) |>
    dplyr::summarise(
      TP = sum(.data$detected & .data$expected_detected, na.rm = TRUE),
      FP = sum(.data$detected & !.data$expected_detected, na.rm = TRUE),
      FN = sum(!.data$detected & .data$expected_detected, na.rm = TRUE),
      precision = dplyr::if_else(.data$TP + .data$FP > 0, .data$TP / (.data$TP + .data$FP), NA_real_),
      recall = dplyr::if_else(.data$TP + .data$FN > 0, .data$TP / (.data$TP + .data$FN), NA_real_),
      F1 = dplyr::if_else(!is.na(.data$precision) & !is.na(.data$recall) & (.data$precision + .data$recall) > 0,
                          2 * .data$precision * .data$recall / (.data$precision + .data$recall), NA_real_),
      false_discovery_rate = dplyr::if_else(.data$TP + .data$FP > 0, .data$FP / (.data$TP + .data$FP), NA_real_),
      Bray_Curtis = calc_bray(.data$relative_abundance, .data$expected_abundance),
      R2 = calc_r2(.data$relative_abundance, .data$expected_abundance),
      .groups = "drop"
    ) |>
    dplyr::mutate(detection_threshold = threshold_label)
}

make_rf_importance <- function(meta_df, outcome_col) {
  predictors <- c("DNA_kit", "BLAST", "Paired", "Trimm", "matrix_type")
  active_predictors <- predictors[purrr::map_lgl(predictors, ~ dplyr::n_distinct(meta_df[[.x]], na.rm = TRUE) > 1)]

  rf_df <- meta_df |>
    dplyr::filter(!is.na(.data[[outcome_col]])) |>
    dplyr::select(dplyr::all_of(c(outcome_col, active_predictors))) |>
    tidyr::drop_na() |>
    dplyr::mutate(dplyr::across(dplyr::all_of(active_predictors), as.factor))

  if (length(active_predictors) == 0 || nrow(rf_df) < 10) {
    return(list(model = NULL, importance = tibble::tibble()))
  }

  set.seed(123)
  rf_model <- randomForest::randomForest(
    formula = stats::as.formula(paste(outcome_col, "~", paste(active_predictors, collapse = " + "))),
    data = rf_df,
    importance = TRUE
  )

  rf_imp <- randomForest::importance(rf_model, type = 1) |>
    as.data.frame() |>
    tibble::rownames_to_column("Variable") |>
    dplyr::rename(IncMSE = `%IncMSE`) |>
    dplyr::mutate(outcome = outcome_col)

  list(model = rf_model, importance = rf_imp)
}

plot_box_jitter <- function(data, x_var, y_var, fill_var, fill_colors, y_lab = NULL) {
  ggplot2::ggplot(
    data,
    ggplot2::aes(x = .data[[x_var]], y = .data[[y_var]], fill = .data[[fill_var]])
  ) +
    ggplot2::geom_boxplot(width = 0.6, color = "black", linewidth = 0.4, outlier.shape = NA, alpha = 0.6) +
    ggplot2::geom_jitter(width = 0.15, size = 2, shape = 21, stroke = 0.4, color = "black", fill = "white") +
    ggplot2::scale_fill_manual(values = fill_colors) +
    ggplot2::labs(x = NULL, y = y_lab %||% y_var) +
    fig_theme(base_size = 12)
}

`%||%` <- function(x, y) if (is.null(x)) y else x
```

# 5. Load and prepare data

```{r prepare-data}
feature_table <- qiime2R::read_qza(paths$table_qza)$data
sample_metadata <- readr::read_tsv(paths$sample_metadata, show_col_types = FALSE) |>
  standardize_metadata()

taxonomy_unite <- read_taxonomy_table(paths$taxonomy_unite_normalized, "UNITE")
taxonomy_blast <- read_taxonomy_table(paths$taxonomy_blast_final, "UNITE+BLAST")
taxonomy_all <- dplyr::bind_rows(taxonomy_unite, taxonomy_blast)

feature_long <- feature_table_to_long(feature_table, unique(taxonomy_all$feature_id)) |>
  dplyr::left_join(sample_metadata, by = "sample_id") |>
  dplyr::mutate(
    pipeline = "ITSdetector",
    preprocess_method = "ITSxpress",
    read_strategy = "PE",
    Paired = "PE",
    Trimm = "ITSxpress"
  )

sample_depth <- feature_long |>
  dplyr::group_by(.data$sample_id) |>
  dplyr::summarise(total_reads = sum(.data$count, na.rm = TRUE), .groups = "drop")

readr::write_tsv(taxonomy_unite, file.path(table_dir, "taxonomy_unite_normalized_used.tsv"))
readr::write_tsv(taxonomy_blast, file.path(table_dir, "taxonomy_unite_blast_used.tsv"))
readr::write_tsv(feature_long, file.path(table_dir, "feature_table_asv_long.tsv"))
```

# 6. Create benchmark long table

```{r benchmark-long}
taxa_abundance <- feature_long |>
  dplyr::select(
    .data$feature_id, .data$sample_id, .data$count, .data$DNA_kit, .data$replicate,
    .data$matrix_type, .data$pipeline, .data$preprocess_method, .data$read_strategy,
    .data$Paired, .data$Trimm
  ) |>
  dplyr::left_join(taxonomy_all, by = "feature_id") |>
  dplyr::mutate(BLAST = dplyr::if_else(.data$taxonomy_method == "UNITE+BLAST", "BLAST", "No_BLAST")) |>
  tidyr::pivot_longer(
    cols = c(.data$Genus, .data$Species),
    names_to = "taxonomic_rank",
    values_to = "taxon"
  ) |>
  dplyr::mutate(
    taxon = dplyr::if_else(is.na(.data$taxon), "Unclassified", .data$taxon),
    taxon = dplyr::if_else(.data$taxon == "", "Unclassified", .data$taxon)
  ) |>
  dplyr::group_by(
    .data$sample_id, .data$DNA_kit, .data$replicate, .data$matrix_type,
    .data$pipeline, .data$preprocess_method, .data$read_strategy,
    .data$taxonomy_method, .data$BLAST, .data$Paired, .data$Trimm,
    .data$taxonomic_rank, .data$taxon
  ) |>
  dplyr::summarise(count = sum(.data$count, na.rm = TRUE), .groups = "drop") |>
  dplyr::left_join(sample_depth, by = "sample_id") |>
  dplyr::mutate(relative_abundance = dplyr::if_else(.data$total_reads > 0, .data$count / .data$total_reads, 0))

sample_method_grid <- taxa_abundance |>
  dplyr::distinct(
    .data$sample_id, .data$DNA_kit, .data$replicate, .data$matrix_type,
    .data$pipeline, .data$preprocess_method, .data$read_strategy,
    .data$taxonomy_method, .data$BLAST, .data$Paired, .data$Trimm
  )

taxon_universe <- dplyr::bind_rows(
  taxa_abundance |> dplyr::distinct(.data$taxonomic_rank, .data$taxon),
  expected_truth |> dplyr::distinct(.data$taxonomic_rank, .data$taxon)
) |>
  dplyr::distinct(.data$taxonomic_rank, .data$taxon)

benchmark_long <- sample_method_grid |>
  tidyr::crossing(taxon_universe) |>
  dplyr::left_join(expected_truth, by = c("taxonomic_rank", "taxon")) |>
  dplyr::mutate(
    expected_abundance = dplyr::case_when(
      .data$matrix_type %in% c("mock_only", "mock_plus_stool") & !is.na(.data$expected_abundance) ~ .data$expected_abundance,
      TRUE ~ 0
    )
  ) |>
  dplyr::left_join(
    taxa_abundance,
    by = c(
      "sample_id", "DNA_kit", "replicate", "matrix_type", "pipeline", "preprocess_method",
      "read_strategy", "taxonomy_method", "BLAST", "Paired", "Trimm", "taxonomic_rank", "taxon"
    )
  ) |>
  dplyr::mutate(
    count = tidyr::replace_na(.data$count, 0),
    total_reads = tidyr::replace_na(.data$total_reads, 0),
    relative_abundance = tidyr::replace_na(.data$relative_abundance, 0)
  ) |>
  dplyr::select(
    .data$sample_id, .data$DNA_kit, .data$replicate, .data$matrix_type,
    .data$pipeline, .data$preprocess_method, .data$read_strategy,
    .data$taxonomy_method, .data$BLAST, .data$Paired, .data$Trimm,
    .data$taxonomic_rank, .data$taxon, .data$count, .data$total_reads,
    .data$relative_abundance, .data$expected_abundance
  )

benchmark_long_species <- benchmark_long |> dplyr::filter(.data$taxonomic_rank == "Species")
benchmark_long_genus <- benchmark_long |> dplyr::filter(.data$taxonomic_rank == "Genus")

readr::write_tsv(benchmark_long, file.path(table_dir, "benchmark_long_genus_species.tsv"))
readr::write_tsv(benchmark_long_species, file.path(table_dir, "benchmark_long_species.tsv"))
readr::write_tsv(benchmark_long_genus, file.path(table_dir, "benchmark_long_genus.tsv"))
```

# 7. Figure 1: Species-level taxa plot

```{r taxa-plot, fig.width=15, fig.height=8}
taxa_plot_df <- benchmark_long_species |>
  dplyr::mutate(
    taxon_plot = dplyr::if_else(.data$taxon %in% expected_species, .data$taxon, "Other"),
    taxon_plot = factor(.data$taxon_plot, levels = c(expected_species, "Other")),
    taxonomy_method = factor(.data$taxonomy_method, levels = c("UNITE", "UNITE+BLAST")),
    DNA_kit = factor(.data$DNA_kit, levels = c("DNeasy", "QIAamp", "Zymo"))
  ) |>
  dplyr::group_by(
    .data$sample_id, .data$DNA_kit, .data$replicate, .data$matrix_type,
    .data$taxonomy_method, .data$taxon_plot
  ) |>
  dplyr::summarise(relative_abundance = sum(.data$relative_abundance, na.rm = TRUE), .groups = "drop")

sample_order <- taxa_plot_df |>
  dplyr::distinct(.data$matrix_type, .data$DNA_kit, .data$replicate, .data$sample_id) |>
  dplyr::arrange(.data$matrix_type, .data$DNA_kit, .data$replicate, .data$sample_id) |>
  dplyr::pull(.data$sample_id)

taxa_plot_df <- taxa_plot_df |>
  dplyr::mutate(sample_id = factor(.data$sample_id, levels = unique(sample_order)))

readr::write_tsv(taxa_plot_df, file.path(table_dir, "figure1_species_taxa_plot_data.tsv"))

p_taxa <- ggplot2::ggplot(
  taxa_plot_df,
  ggplot2::aes(x = .data$sample_id, y = .data$relative_abundance, fill = .data$taxon_plot)
) +
  ggplot2::geom_col(width = 0.85, color = NA) +
  ggh4x::facet_nested(taxonomy_method ~ matrix_type + DNA_kit, scales = "free_x", space = "free_x") +
  ggplot2::scale_fill_manual(values = stats::setNames(tax_col, c(expected_species, "Other")), drop = FALSE) +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1), expand = c(0, 0)) +
  ggplot2::labs(x = NULL, y = "Relative abundance", fill = "Species") +
  fig_theme(base_size = 12) +
  ggplot2::theme(axis.text.x.bottom = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7))

p_taxa

ggplot2::ggsave(file.path(figure_dir, "figure1_species_taxa_plot.png"), p_taxa, width = 16, height = 8, dpi = 300)
ggplot2::ggsave(file.path(figure_dir, "figure1_species_taxa_plot.pdf"), p_taxa, width = 16, height = 8)
```

# 8. Figure 2: Ground-truth accuracy metrics

```{r metric-calc-plot, fig.width=13, fig.height=9}
metrics_main <- calc_metrics(benchmark_long, threshold_value = 0, threshold_label = "count_gt_0")
metrics_sensitivity <- calc_metrics(benchmark_long, threshold_value = 0.001, threshold_label = "relative_abundance_ge_0.001")

readr::write_tsv(metrics_main, file.path(table_dir, "benchmark_metrics_main_count_gt_0.tsv"))
readr::write_tsv(metrics_sensitivity, file.path(table_dir, "benchmark_metrics_sensitivity_ra_ge_0.001.tsv"))

meta2 <- metrics_main |>
  dplyr::filter(.data$taxonomic_rank == "Species") |>
  dplyr::mutate(
    DNA_kit = factor(.data$DNA_kit, levels = c("DNeasy", "QIAamp", "Zymo")),
    BLAST = factor(.data$BLAST, levels = c("No_BLAST", "BLAST")),
    taxonomy_method = factor(.data$taxonomy_method, levels = c("UNITE", "UNITE+BLAST"))
  )

metric_plot_df <- meta2 |>
  dplyr::select(
    .data$sample_id, .data$DNA_kit, .data$replicate, .data$matrix_type,
    .data$taxonomy_method, .data$BLAST, .data$F1, .data$R2, .data$Bray_Curtis,
    .data$precision, .data$recall, .data$false_discovery_rate
  ) |>
  tidyr::pivot_longer(
    cols = c(.data$F1, .data$R2, .data$Bray_Curtis, .data$precision, .data$recall, .data$false_discovery_rate),
    names_to = "metric",
    values_to = "value"
  ) |>
  dplyr::mutate(
    metric = dplyr::recode(
      .data$metric,
      "F1" = "F1-score",
      "R2" = "R²",
      "Bray_Curtis" = "Bray–Curtis distance",
      "precision" = "Precision",
      "recall" = "Recall",
      "false_discovery_rate" = "False discovery rate"
    ),
    metric = factor(
      .data$metric,
      levels = c("F1-score", "R²", "Bray–Curtis distance", "Precision", "Recall", "False discovery rate")
    )
  )

readr::write_tsv(meta2, file.path(table_dir, "figure2_species_metric_table.tsv"))
readr::write_tsv(metric_plot_df, file.path(table_dir, "figure2_species_metric_plot_data.tsv"))

p_metrics <- plot_box_jitter(
  data = metric_plot_df,
  x_var = "taxonomy_method",
  y_var = "value",
  fill_var = "taxonomy_method",
  fill_colors = c("UNITE" = met_col[1], "UNITE+BLAST" = met_col[2]),
  y_lab = "Metric value"
) +
  ggplot2::facet_grid(metric ~ matrix_type + DNA_kit, scales = "free_y") +
  ggplot2::theme(strip.text.y = ggplot2::element_text(angle = 0, face = "bold"))

p_metrics

ggplot2::ggsave(file.path(figure_dir, "figure2_ground_truth_accuracy_metrics.png"), p_metrics, width = 14, height = 9, dpi = 300)
ggplot2::ggsave(file.path(figure_dir, "figure2_ground_truth_accuracy_metrics.pdf"), p_metrics, width = 14, height = 9)
```

# 9. Figure 3: Random forest importance and PERMANOVA R²

```{r rf-permanova, fig.width=11, fig.height=7}
rf_r2_out <- make_rf_importance(meta2, "R2")
rf_f1_out <- make_rf_importance(meta2, "F1")

rf_r2 <- rf_r2_out$model
rf_f1 <- rf_f1_out$model
importance_r2.1 <- rf_r2_out$importance
importance_f1.1 <- rf_f1_out$importance

importance_all <- dplyr::bind_rows(
  importance_r2.1 |> dplyr::mutate(outcome = "R²"),
  importance_f1.1 |> dplyr::mutate(outcome = "F1-score")
)

readr::write_tsv(importance_r2.1, file.path(table_dir, "rf_importance_R2.tsv"))
readr::write_tsv(importance_f1.1, file.path(table_dir, "rf_importance_F1.tsv"))
readr::write_tsv(importance_all, file.path(table_dir, "rf_importance_combined.tsv"))

plot_effect <- function(df, title_text) {
  ggplot2::ggplot(df, ggplot2::aes(x = stats::reorder(.data$Variable, .data$IncMSE), y = .data$IncMSE)) +
    ggplot2::geom_col(fill = "steelblue") +
    ggplot2::coord_flip() +
    fig_theme(base_size = 13) +
    ggplot2::labs(x = NULL, y = "Permutation importance (%IncMSE)", title = title_text)
}

p_rf_r2 <- plot_effect(importance_r2.1, "R²")
p_rf_f1 <- plot_effect(importance_f1.1, "F1-score")
p_rf <- p_rf_r2 / p_rf_f1

community_wide <- benchmark_long_species |>
  dplyr::filter(.data$matrix_type %in% c("mock_only", "mock_plus_stool")) |>
  dplyr::mutate(sample_method_id = paste(.data$sample_id, .data$taxonomy_method, sep = "__")) |>
  dplyr::select(.data$sample_method_id, .data$taxon, .data$relative_abundance) |>
  dplyr::group_by(.data$sample_method_id, .data$taxon) |>
  dplyr::summarise(relative_abundance = sum(.data$relative_abundance, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = .data$taxon, values_from = .data$relative_abundance, values_fill = 0)

perma_meta <- benchmark_long_species |>
  dplyr::filter(.data$matrix_type %in% c("mock_only", "mock_plus_stool")) |>
  dplyr::mutate(sample_method_id = paste(.data$sample_id, .data$taxonomy_method, sep = "__")) |>
  dplyr::distinct(.data$sample_method_id, .data$sample_id, .data$DNA_kit, .data$BLAST, .data$Paired, .data$Trimm, .data$matrix_type)

community_mat <- community_wide |>
  dplyr::arrange(.data$sample_method_id) |>
  tibble::column_to_rownames("sample_method_id") |>
  as.matrix()

perma_meta <- perma_meta |>
  dplyr::filter(.data$sample_method_id %in% rownames(community_mat)) |>
  dplyr::arrange(match(.data$sample_method_id, rownames(community_mat)))

perma_vars <- c("DNA_kit", "BLAST", "Paired", "Trimm", "matrix_type")
perma_vars <- perma_vars[purrr::map_lgl(perma_vars, ~ dplyr::n_distinct(perma_meta[[.x]], na.rm = TRUE) > 1)]

set.seed(123)
dist_matrix <- vegan::vegdist(community_mat, method = "bray")
per_res <- vegan::adonis2(
  formula = stats::as.formula(paste("dist_matrix ~", paste(perma_vars, collapse = " + "))),
  data = perma_meta,
  permutations = 9999,
  by = "margin"
)

perma_out <- per_res |>
  as.data.frame() |>
  tibble::rownames_to_column("Variable") |>
  dplyr::filter(!.data$Variable %in% c("Residual", "Total")) |>
  dplyr::rename(p_value = `Pr(>F)`)

readr::write_tsv(perma_out, file.path(table_dir, "permanova_R2.tsv"))

p_permanova <- ggplot2::ggplot(perma_out, ggplot2::aes(x = stats::reorder(.data$Variable, .data$R2), y = .data$R2)) +
  ggplot2::geom_col(fill = "steelblue") +
  ggplot2::coord_flip() +
  fig_theme(base_size = 13) +
  ggplot2::labs(x = NULL, y = "PERMANOVA R²", title = "PERMANOVA effect size")

p_rf
p_permanova

ggplot2::ggsave(file.path(figure_dir, "figure3A_random_forest_importance.png"), p_rf, width = 9, height = 7, dpi = 300)
ggplot2::ggsave(file.path(figure_dir, "figure3A_random_forest_importance.pdf"), p_rf, width = 9, height = 7)
ggplot2::ggsave(file.path(figure_dir, "figure3B_permanova_R2.png"), p_permanova, width = 7, height = 4, dpi = 300)
ggplot2::ggsave(file.path(figure_dir, "figure3B_permanova_R2.pdf"), p_permanova, width = 7, height = 4)
```

# 10. Combined figure

```{r combined-figure, fig.width=16, fig.height=18}
fig_combined <- p_taxa /
  p_metrics /
  (p_rf | p_permanova) +
  patchwork::plot_layout(heights = c(2.2, 2.2, 1.2)) +
  patchwork::plot_annotation(title = "ITSxpress PE benchmark: UNITE-normalized vs UNITE+BLAST")

fig_combined

ggplot2::ggsave(file.path(figure_dir, "figure_combined_ITSxpress_PE_corrected.png"), fig_combined, width = 18, height = 20, dpi = 300)
ggplot2::ggsave(file.path(figure_dir, "figure_combined_ITSxpress_PE_corrected.pdf"), fig_combined, width = 18, height = 20)
```

# 11. Session information

```{r session-info}
sessionInfo()
```
