---
title: "Figure 4 pilot: ITSxpress PE, taxonomy normalized vs BLAST"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

options(stringsAsFactors = FALSE)
```

# 1. Brief

This Rmd uses one completed ITSdetector result set.

- Abundance table: `dada2/itsxpress/paired/table.qza`
- Pre-BLAST taxonomy: `taxonomy_blast/.../taxonomy_normalized.tsv`
- Post-BLAST taxonomy: `taxonomy_blast/.../taxonomy_blast_qiime.tsv`
- Main metric threshold: observed if `relative_abundance > 0`
- Sensitivity metric threshold: observed if `relative_abundance >= 0.001`
- PBS controls are kept in the taxa plot but excluded from ground-truth metric summaries.

# 2. Paths and packages

```{r paths}
required_pkgs <- c(
  "qiime2R", "readr", "dplyr", "tidyr", "tibble", "stringr",
  "ggplot2", "scales", "vegan", "patchwork"
)

missing_pkgs <- required_pkgs[
  !purrr::map_lgl(required_pkgs, requireNamespace, quietly = TRUE)
]

if (length(missing_pkgs) > 0) {
  stop("Missing R packages: ", paste(missing_pkgs, collapse = ", "))
}

project_dir <- "/data/home2/ksy/260724_GutMyco_Fungal_Mock_Paper"
rproj_dir <- file.path(project_dir, "Rproj")

run_id <- "fig4_q20_PE_itsxpress_unite"
run_dir <- file.path(project_dir, "Output", run_id)

tax_dir <- file.path(
  run_dir,
  "taxonomy_blast", "itsxpress", "paired", "unite_ver10_99"
)

paths <- list(
  table_qza = file.path(run_dir, "dada2", "itsxpress", "paired", "table.qza"),
  repseq_qza = file.path(run_dir, "dada2", "itsxpress", "paired", "repseq.qza"),
  denoising_stats_qza = file.path(run_dir, "dada2", "itsxpress", "paired", "denoising_stats.qza"),
  taxonomy_normalized_tsv = file.path(tax_dir, "taxonomy_normalized.tsv"),
  taxonomy_blast_qiime_tsv = file.path(tax_dir, "taxonomy_blast_qiime.tsv"),
  blast_evidence_tsv = file.path(tax_dir, "taxonomy_blast_evidence.tsv"),
  blast_report_tsv = file.path(tax_dir, "taxonomy_blast_report.tsv"),
  metadata_tsv = file.path(project_dir, "Input", "metadata_mock_stool_and_mock_260724.tsv")
)

out_dir <- file.path(rproj_dir, "Output", "fig4_pilot_itsxpress_PE_normalized_vs_blast")
table_dir <- file.path(out_dir, "tables")
figure_dir <- file.path(out_dir, "figures")

dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

path_check <- tibble::tibble(
  file_key = names(paths),
  path = unlist(paths),
  exists = file.exists(unlist(paths))
)

readr::write_tsv(path_check, file.path(table_dir, "input_path_check.tsv"))
print(path_check)

if (any(!path_check$exists)) {
  stop("Missing required input files. See input_path_check.tsv")
}
```

# 3. Load input files

```{r load-inputs}
feature_table <- qiime2R::read_qza(paths$table_qza)$data
repseq <- qiime2R::read_qza(paths$repseq_qza)$data
denoising_stats <- qiime2R::read_qza(paths$denoising_stats_qza)$data

taxonomy_normalized <- readr::read_tsv(
  paths$taxonomy_normalized_tsv,
  show_col_types = FALSE
)

taxonomy_blast <- readr::read_tsv(
  paths$taxonomy_blast_qiime_tsv,
  show_col_types = FALSE
)

blast_evidence <- readr::read_tsv(
  paths$blast_evidence_tsv,
  show_col_types = FALSE
)

blast_report <- readr::read_tsv(
  paths$blast_report_tsv,
  show_col_types = FALSE
)

sample_metadata <- readr::read_tsv(
  paths$metadata_tsv,
  show_col_types = FALSE
)
```

# 4. Minimal helper functions

```{r helper-functions}
clean_taxon <- function(x) {
  x |>
    as.character() |>
    stringr::str_replace("^.__", "") |>
    stringr::str_replace_all(" ", "_") |>
    stringr::str_replace_all("^_+|_+$", "") |>
    dplyr::na_if("")
}

normalize_sample_id <- function(x) {
  x |>
    as.character() |>
    stringr::str_replace("^X", "")
}

standardize_metadata <- function(metadata_df) {
  metadata_df |>
    dplyr::rename(
      sample_id = dplyr::any_of("sample"),
      kit = dplyr::any_of("kit_code")
    ) |>
    dplyr::mutate(
      sample_id = normalize_sample_id(.data$sample_id),
      kit = as.character(.data$kit),
      replicate = as.character(.data$replicate),
      matrix_type = as.character(.data$matrix_type)
    ) |>
    dplyr::select(
      .data$sample_id,
      .data$kit,
      .data$replicate,
      .data$matrix_type,
      dplyr::everything()
    ) |>
    dplyr::distinct(.data$sample_id, .keep_all = TRUE)
}

standardize_taxonomy <- function(tax_df, taxonomy_method_label) {
  tax <- tax_df |>
    dplyr::rename(
      feature_id = dplyr::any_of(c("Feature ID", "Feature.ID", "feature_id")),
      taxon_string = dplyr::any_of(c("Taxon", "taxonomy", "taxon_string"))
    )

  if (!"feature_id" %in% colnames(tax)) {
    stop("No feature ID column found in taxonomy table.")
  }

  if ("Genus" %in% colnames(tax) && "Species" %in% colnames(tax)) {
    tax |>
      dplyr::transmute(
        feature_id = .data$feature_id,
        taxon_string = .data$taxon_string,
        Confidence = .data$Confidence,
        taxonomy_method = taxonomy_method_label,
        Genus = clean_taxon(.data$Genus),
        Species = clean_taxon(.data$Species)
      ) |>
      dplyr::mutate(
        Genus = tidyr::replace_na(.data$Genus, "Unclassified"),
        Species = tidyr::replace_na(.data$Species, "Unclassified")
      )
  } else {
    split_tax <- stringr::str_split_fixed(tax$taxon_string, ";", 7)

    tibble::as_tibble(split_tax) |>
      dplyr::rename(
        Kingdom = V1,
        Phylum = V2,
        Class = V3,
        Order = V4,
        Family = V5,
        Genus = V6,
        Species = V7
      ) |>
      dplyr::mutate(
        dplyr::across(
          .cols = dplyr::everything(),
          .fns = clean_taxon
        )
      ) |>
      dplyr::bind_cols(
        tax |>
          dplyr::select(.data$feature_id, .data$taxon_string, dplyr::any_of("Confidence"))
      ) |>
      dplyr::transmute(
        feature_id = .data$feature_id,
        taxon_string = .data$taxon_string,
        Confidence = .data$Confidence,
        taxonomy_method = taxonomy_method_label,
        Genus = tidyr::replace_na(.data$Genus, "Unclassified"),
        Species = tidyr::replace_na(.data$Species, "Unclassified")
      )
  }
}

make_feature_table_long <- function(feature_table_obj, feature_ids) {
  ft <- as.data.frame(feature_table_obj)

  if (sum(colnames(ft) %in% feature_ids) > sum(rownames(ft) %in% feature_ids)) {
    ft <- t(ft) |> as.data.frame()
  }

  ft |>
    tibble::rownames_to_column("feature_id") |>
    tidyr::pivot_longer(
      cols = - .data$feature_id,
      names_to = "sample_id",
      values_to = "count"
    ) |>
    dplyr::mutate(
      sample_id = normalize_sample_id(.data$sample_id),
      count = as.numeric(.data$count)
    )
}

calculate_bray <- function(observed, expected) {
  if (sum(observed, expected, na.rm = TRUE) == 0) {
    return(NA_real_)
  }

  as.numeric(
    vegan::vegdist(
      rbind(observed, expected),
      method = "bray"
    )[1]
  )
}

calculate_r2 <- function(observed, expected) {
  if (length(unique(expected)) < 2 || sum(observed, na.rm = TRUE) == 0) {
    return(NA_real_)
  }

  summary(stats::lm(observed ~ expected))$r.squared
}

calculate_metrics <- function(long_df, threshold_value, threshold_label) {
  long_df |>
    dplyr::filter(.data$matrix_type %in% c("mock_only", "mock_plus_stool")) |>
    dplyr::mutate(
      detected = .data$relative_abundance > threshold_value,
      expected_detected = .data$expected_abundance > 0
    ) |>
    dplyr::group_by(
      .data$sample_id,
      .data$kit,
      .data$replicate,
      .data$matrix_type,
      .data$pipeline,
      .data$preprocess_method,
      .data$read_strategy,
      .data$taxonomy_method,
      .data$taxonomic_rank
    ) |>
    dplyr::summarise(
      TP = sum(.data$detected & .data$expected_detected, na.rm = TRUE),
      FP = sum(.data$detected & !.data$expected_detected, na.rm = TRUE),
      FN = sum(!.data$detected & .data$expected_detected, na.rm = TRUE),
      precision = dplyr::if_else(TP + FP > 0, TP / (TP + FP), NA_real_),
      recall = dplyr::if_else(TP + FN > 0, TP / (TP + FN), NA_real_),
      F1_score = dplyr::if_else(
        !is.na(.data$precision) &
          !is.na(.data$recall) &
          (.data$precision + .data$recall) > 0,
        2 * .data$precision * .data$recall / (.data$precision + .data$recall),
        NA_real_
      ),
      false_discovery_rate = dplyr::if_else(TP + FP > 0, FP / (TP + FP), NA_real_),
      Bray_Curtis_distance = calculate_bray(.data$relative_abundance, .data$expected_abundance),
      R2 = calculate_r2(.data$relative_abundance, .data$expected_abundance),
      .groups = "drop"
    ) |>
    dplyr::mutate(detection_threshold = threshold_label)
}
```

# 5. Ground truth

```{r ground-truth}
mock_species <- c(
  "Aspergillus_fumigatus",
  "Cryptococcus_neoformans",
  "Trichophyton_interdigitale",
  "Penicillium_chrysogenum",
  "Fusarium_keratoplasticum",
  "Candida_albicans",
  "Nakaseomyces_glabratus",
  "Malassezia_globosa",
  "Saccharomyces_cerevisiae",
  "Cutaneotrichosporon_dermatis"
)

mock_genus <- c(
  "Aspergillus",
  "Cryptococcus",
  "Trichophyton",
  "Penicillium",
  "Fusarium",
  "Candida",
  "Nakaseomyces",
  "Malassezia",
  "Saccharomyces",
  "Cutaneotrichosporon"
)

expected_truth <- dplyr::bind_rows(
  tibble::tibble(
    taxonomic_rank = "Species",
    taxon = mock_species,
    expected_abundance = 0.1
  ),
  tibble::tibble(
    taxonomic_rank = "Genus",
    taxon = mock_genus,
    expected_abundance = 0.1
  )
)

readr::write_tsv(
  expected_truth,
  file.path(table_dir, "expected_truth.tsv")
)
```

# 6. Build long-format tables

```{r build-long-table}
sample_metadata2 <- standardize_metadata(sample_metadata)

taxonomy_before <- standardize_taxonomy(
  tax_df = taxonomy_normalized,
  taxonomy_method_label = "UNITE-normalized"
)

taxonomy_after <- standardize_taxonomy(
  tax_df = taxonomy_blast,
  taxonomy_method_label = "UNITE+BLAST"
)

taxonomy_all <- dplyr::bind_rows(
  taxonomy_before,
  taxonomy_after
)

feature_table_long <- make_feature_table_long(
  feature_table_obj = feature_table,
  feature_ids = unique(taxonomy_all$feature_id)
)

sample_depth <- feature_table_long |>
  dplyr::group_by(.data$sample_id) |>
  dplyr::summarise(
    total_reads = sum(.data$count, na.rm = TRUE),
    .groups = "drop"
  )

observed_long <- feature_table_long |>
  dplyr::left_join(sample_metadata2, by = "sample_id") |>
  dplyr::mutate(
    pipeline = "ITSdetector",
    preprocess_method = "ITSxpress",
    read_strategy = "PE"
  ) |>
  dplyr::left_join(taxonomy_all, by = "feature_id") |>
  tidyr::pivot_longer(
    cols = c(.data$Genus, .data$Species),
    names_to = "taxonomic_rank",
    values_to = "taxon"
  ) |>
  dplyr::mutate(
    taxon = tidyr::replace_na(.data$taxon, "Unclassified")
  ) |>
  dplyr::group_by(
    .data$sample_id,
    .data$kit,
    .data$replicate,
    .data$matrix_type,
    .data$pipeline,
    .data$preprocess_method,
    .data$read_strategy,
    .data$taxonomy_method,
    .data$taxonomic_rank,
    .data$taxon
  ) |>
  dplyr::summarise(
    count = sum(.data$count, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::left_join(sample_depth, by = "sample_id") |>
  dplyr::mutate(
    relative_abundance = dplyr::if_else(
      .data$total_reads > 0,
      .data$count / .data$total_reads,
      0
    )
  )

sample_method_grid <- observed_long |>
  dplyr::distinct(
    .data$sample_id,
    .data$kit,
    .data$replicate,
    .data$matrix_type,
    .data$pipeline,
    .data$preprocess_method,
    .data$read_strategy,
    .data$taxonomy_method
  )

taxon_universe <- observed_long |>
  dplyr::distinct(.data$taxonomic_rank, .data$taxon) |>
  dplyr::bind_rows(
    expected_truth |>
      dplyr::select(.data$taxonomic_rank, .data$taxon)
  ) |>
  dplyr::distinct(.data$taxonomic_rank, .data$taxon)

benchmark_long <- sample_method_grid |>
  tidyr::crossing(taxon_universe) |>
  dplyr::left_join(
    expected_truth,
    by = c("taxonomic_rank", "taxon")
  ) |>
  dplyr::mutate(
    expected_abundance = dplyr::case_when(
      .data$matrix_type %in% c("mock_only", "mock_plus_stool") &
        !is.na(.data$expected_abundance) ~ .data$expected_abundance,
      TRUE ~ 0
    )
  ) |>
  dplyr::left_join(
    observed_long,
    by = c(
      "sample_id", "kit", "replicate", "matrix_type",
      "pipeline", "preprocess_method", "read_strategy",
      "taxonomy_method", "taxonomic_rank", "taxon"
    )
  ) |>
  dplyr::mutate(
    count = tidyr::replace_na(.data$count, 0),
    total_reads = tidyr::replace_na(.data$total_reads, 0),
    relative_abundance = tidyr::replace_na(.data$relative_abundance, 0)
  ) |>
  dplyr::select(
    .data$sample_id,
    .data$kit,
    .data$replicate,
    .data$matrix_type,
    .data$pipeline,
    .data$preprocess_method,
    .data$read_strategy,
    .data$taxonomy_method,
    .data$taxonomic_rank,
    .data$taxon,
    .data$count,
    .data$total_reads,
    .data$relative_abundance,
    .data$expected_abundance
  )

readr::write_tsv(
  taxonomy_before,
  file.path(table_dir, "taxonomy_before_unite_normalized.tsv")
)

readr::write_tsv(
  taxonomy_after,
  file.path(table_dir, "taxonomy_after_unite_blast.tsv")
)

readr::write_tsv(
  observed_long,
  file.path(table_dir, "observed_long.tsv")
)

readr::write_tsv(
  benchmark_long,
  file.path(table_dir, "benchmark_long.tsv")
)
```

# 7. Metrics

```{r metrics}
metrics_main <- calculate_metrics(
  long_df = benchmark_long,
  threshold_value = 0,
  threshold_label = "relative_abundance_gt_0"
)

metrics_sensitivity <- calculate_metrics(
  long_df = benchmark_long,
  threshold_value = 0.001,
  threshold_label = "relative_abundance_ge_0.001"
)

metrics_species_main <- metrics_main |>
  dplyr::filter(.data$taxonomic_rank == "Species")

readr::write_tsv(
  metrics_main,
  file.path(table_dir, "metrics_main.tsv")
)

readr::write_tsv(
  metrics_sensitivity,
  file.path(table_dir, "metrics_sensitivity.tsv")
)

readr::write_tsv(
  metrics_species_main,
  file.path(table_dir, "metrics_species_main.tsv")
)
```

# 8. Figure 1. Species taxa plot

```{r taxa-plot, fig.width=14, fig.height=8}
taxa_plot_df <- benchmark_long |>
  dplyr::filter(.data$taxonomic_rank == "Species") |>
  dplyr::mutate(
    taxon_plot = dplyr::if_else(.data$taxon %in% mock_species, .data$taxon, "Other"),
    taxon_plot = factor(.data$taxon_plot, levels = c(mock_species, "Other")),
    taxonomy_method = factor(.data$taxonomy_method, levels = c("UNITE-normalized", "UNITE+BLAST"))
  ) |>
  dplyr::group_by(
    .data$sample_id,
    .data$kit,
    .data$replicate,
    .data$matrix_type,
    .data$taxonomy_method,
    .data$taxon_plot
  ) |>
  dplyr::summarise(
    relative_abundance = sum(.data$relative_abundance, na.rm = TRUE),
    .groups = "drop"
  )

sample_order <- taxa_plot_df |>
  dplyr::distinct(.data$matrix_type, .data$kit, .data$replicate, .data$sample_id) |>
  dplyr::arrange(.data$matrix_type, .data$kit, .data$replicate, .data$sample_id) |>
  dplyr::pull(.data$sample_id)

taxa_plot_df <- taxa_plot_df |>
  dplyr::mutate(sample_id = factor(.data$sample_id, levels = unique(sample_order)))

facet_layer <- if (requireNamespace("ggh4x", quietly = TRUE)) {
  ggh4x::facet_nested(
    taxonomy_method ~ matrix_type + kit,
    scales = "free_x",
    space = "free_x"
  )
} else {
  ggplot2::facet_grid(
    taxonomy_method ~ matrix_type + kit,
    scales = "free_x",
    space = "free_x"
  )
}

p_taxa <- ggplot2::ggplot(
  taxa_plot_df,
  ggplot2::aes(
    x = .data$sample_id,
    y = .data$relative_abundance,
    fill = .data$taxon_plot
  )
) +
  ggplot2::geom_col(width = 0.85) +
  facet_layer +
  ggplot2::scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = c(0, 0)
  ) +
  ggplot2::labs(
    x = NULL,
    y = "Relative abundance",
    fill = "Species",
    title = "Species-level composition"
  ) +
  ggplot2::theme_bw(base_size = 10) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
    legend.position = "right",
    panel.grid.major.x = ggplot2::element_blank()
  )

p_taxa

readr::write_tsv(
  taxa_plot_df,
  file.path(table_dir, "figure1_taxa_plot_species_data.tsv")
)

ggplot2::ggsave(
  file.path(figure_dir, "figure1_species_taxa_plot.png"),
  p_taxa,
  width = 16,
  height = 8,
  dpi = 300
)

ggplot2::ggsave(
  file.path(figure_dir, "figure1_species_taxa_plot.pdf"),
  p_taxa,
  width = 16,
  height = 8
)
```

# 9. Figure 2. Ground-truth metrics

```{r metric-plot, fig.width=12, fig.height=8}
metric_plot_df <- metrics_species_main |>
  dplyr::select(
    .data$sample_id,
    .data$kit,
    .data$replicate,
    .data$matrix_type,
    .data$taxonomy_method,
    .data$F1_score,
    .data$R2,
    .data$Bray_Curtis_distance,
    .data$precision,
    .data$recall,
    .data$false_discovery_rate
  ) |>
  tidyr::pivot_longer(
    cols = c(
      .data$F1_score,
      .data$R2,
      .data$Bray_Curtis_distance,
      .data$precision,
      .data$recall,
      .data$false_discovery_rate
    ),
    names_to = "metric",
    values_to = "value"
  ) |>
  dplyr::mutate(
    metric = dplyr::recode(
      .data$metric,
      F1_score = "F1-score",
      R2 = "R²",
      Bray_Curtis_distance = "Bray–Curtis distance",
      precision = "Precision",
      recall = "Recall",
      false_discovery_rate = "False discovery rate"
    ),
    metric = factor(
      .data$metric,
      levels = c(
        "F1-score", "R²", "Bray–Curtis distance",
        "Precision", "Recall", "False discovery rate"
      )
    )
  )

metric_summary <- metric_plot_df |>
  dplyr::group_by(
    .data$matrix_type,
    .data$kit,
    .data$taxonomy_method,
    .data$metric
  ) |>
  dplyr::summarise(
    mean_value = mean(.data$value, na.rm = TRUE),
    se_value = stats::sd(.data$value, na.rm = TRUE) / sqrt(sum(!is.na(.data$value))),
    .groups = "drop"
  )

p_metrics <- ggplot2::ggplot(
  metric_summary,
  ggplot2::aes(
    x = .data$taxonomy_method,
    y = .data$mean_value,
    fill = .data$taxonomy_method
  )
) +
  ggplot2::geom_col(width = 0.65, color = "black", linewidth = 0.25) +
  ggplot2::geom_errorbar(
    ggplot2::aes(
      ymin = .data$mean_value - .data$se_value,
      ymax = .data$mean_value + .data$se_value
    ),
    width = 0.2,
    linewidth = 0.25
  ) +
  ggplot2::facet_grid(
    metric ~ matrix_type + kit,
    scales = "free_y"
  ) +
  ggplot2::labs(
    x = NULL,
    y = "Metric value",
    fill = "Taxonomy method",
    title = "Ground-truth accuracy"
  ) +
  ggplot2::theme_bw(base_size = 9) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    strip.text.y = ggplot2::element_text(angle = 0)
  )

p_metrics

readr::write_tsv(
  metric_plot_df,
  file.path(table_dir, "figure2_metrics_sample_level.tsv")
)

readr::write_tsv(
  metric_summary,
  file.path(table_dir, "figure2_metrics_summary.tsv")
)

ggplot2::ggsave(
  file.path(figure_dir, "figure2_ground_truth_metrics.png"),
  p_metrics,
  width = 14,
  height = 9,
  dpi = 300
)

ggplot2::ggsave(
  file.path(figure_dir, "figure2_ground_truth_metrics.pdf"),
  p_metrics,
  width = 14,
  height = 9
)
```

# 10. Figure 3. PERMANOVA R²

```{r permanova, fig.width=7, fig.height=4}
community_df <- benchmark_long |>
  dplyr::filter(
    .data$taxonomic_rank == "Species",
    .data$matrix_type %in% c("mock_only", "mock_plus_stool")
  ) |>
  dplyr::mutate(
    sample_method_id = paste(.data$sample_id, .data$taxonomy_method, sep = "__")
  )

community_wide <- community_df |>
  dplyr::select(
    .data$sample_method_id,
    .data$taxon,
    .data$relative_abundance
  ) |>
  tidyr::pivot_wider(
    names_from = .data$taxon,
    values_from = .data$relative_abundance,
    values_fill = 0
  )

community_meta <- community_df |>
  dplyr::distinct(
    .data$sample_method_id,
    .data$kit,
    .data$matrix_type,
    .data$taxonomy_method
  ) |>
  dplyr::arrange(.data$sample_method_id)

community_mat <- community_wide |>
  dplyr::arrange(.data$sample_method_id) |>
  tibble::column_to_rownames("sample_method_id") |>
  as.matrix()

community_meta <- community_meta |>
  dplyr::filter(.data$sample_method_id %in% rownames(community_mat)) |>
  dplyr::arrange(match(.data$sample_method_id, rownames(community_mat)))

set.seed(123)

permanova_res <- vegan::adonis2(
  vegan::vegdist(community_mat, method = "bray") ~ taxonomy_method + kit + matrix_type,
  data = community_meta,
  permutations = 9999,
  by = "margin"
)

permanova_table <- permanova_res |>
  as.data.frame() |>
  tibble::rownames_to_column("variable") |>
  dplyr::filter(!.data$variable %in% c("Residual", "Total")) |>
  dplyr::rename(p_value = `Pr(>F)`) |>
  dplyr::select(
    .data$variable,
    .data$Df,
    .data$SumOfSqs,
    .data$R2,
    .data$F,
    .data$p_value
  )

p_permanova <- ggplot2::ggplot(
  permanova_table,
  ggplot2::aes(
    x = .data$R2,
    y = stats::reorder(.data$variable, .data$R2)
  )
) +
  ggplot2::geom_col(width = 0.65) +
  ggplot2::labs(
    x = "PERMANOVA R²",
    y = NULL,
    title = "Effect size on species-level composition"
  ) +
  ggplot2::theme_bw(base_size = 11)

p_permanova

readr::write_tsv(
  permanova_table,
  file.path(table_dir, "figure3_permanova_R2.tsv")
)

ggplot2::ggsave(
  file.path(figure_dir, "figure3_permanova_R2.png"),
  p_permanova,
  width = 7,
  height = 4,
  dpi = 300
)

ggplot2::ggsave(
  file.path(figure_dir, "figure3_permanova_R2.pdf"),
  p_permanova,
  width = 7,
  height = 4
)
```

# 11. Combined figure

```{r combined-figure, fig.width=16, fig.height=18}
combined_figure <- patchwork::wrap_plots(
  p_taxa,
  p_metrics,
  p_permanova,
  ncol = 1,
  heights = c(2.2, 2.2, 1)
) +
  patchwork::plot_annotation(
    title = "ITSxpress PE pilot: UNITE-normalized vs UNITE+BLAST"
  )

combined_figure

ggplot2::ggsave(
  file.path(figure_dir, "figure_combined.png"),
  combined_figure,
  width = 18,
  height = 20,
  dpi = 300
)

ggplot2::ggsave(
  file.path(figure_dir, "figure_combined.pdf"),
  combined_figure,
  width = 18,
  height = 20
)
```

# 12. Session info

```{r session-info}
sessionInfo()
```
